Cntnap2 loss drives striatal neuron hyperexcitability and behavioral inflexibility

Version 1

Abstract

Abstract

Autism spectrum disorder (ASD) is a neurodevelopmental disorder characterized by two major diagnostic criteria - persistent deficits in social communication and interaction, and the presence of restricted, repetitive patterns of behavior (RRBs). Evidence from both human and animal model studies of ASD suggest that alteration of striatal circuits, which mediate motor learning, action selection, and habit formation, may contribute to the manifestation of RRBs. CNTNAP2 is a syndromic ASD risk gene, and loss of function of Cntnap2 in mice is associated with RRBs. How loss of Cntnap2 impacts striatal neuron function is largely unknown. In this study, we utilized Cntnap2-/- mice to test whether altered striatal neuron activity contributes to aberrant motor behaviors relevant to ASD. We find that Cntnap2-/- mice exhibit increased cortical drive of striatal projection neurons (SPNs), with the most pronounced effects in direct pathway SPNs. This enhanced drive is likely due to increased intrinsic excitability of SPNs, which make them more responsive to cortical inputs. We also find that Cntnap2-/- mice exhibit spontaneous repetitive behaviors, increased motor routine learning, and cognitive inflexibility. Increased corticostriatal drive, in particular of the direct pathway, may contribute to the acquisition of repetitive, inflexible behaviors in Cntnap2 mice.

Introduction

Autism spectrum disorder (ASD) is a neurodevelopmental disorder characterized by alterations in social communication and interaction, as well as the presentation of restricted, repetitive behaviors (APA, 2022). Given that ASD has high heritability (Sandin et al., 2017), much work has been done in the last 30 years to identify genes that confer risk of developing ASD (De Rubeis et al., 2014; Iossifov et al., 2014; Sanders et al., 2015). Through this, hundreds of high-confidence risk genes have been identified, varying greatly in the proteins for which they code (Satterstrom et al., 2020). These include transcriptional and translational regulators, ion channels, receptors, cell adhesion molecules, and others (De Rubeis et al., 2014; Delorme et al., 2013; Ebert & Greenberg, 2013). Recent work has focused on identifying brain regions and circuits that may be commonly affected by ASD-related mutations. The basal ganglia, in particular the striatum, represents one such commonly altered brain region and prior studies have demonstrated changes in striatal circuit function and striatum-associated behaviors in mice with mutations in ASD risk genes (Benthall et al., 2021; Fuccillo, 2016; Peca et al., 2011; Peixoto et al., 2016; Platt et al., 2017; Rothwell et al., 2014; Wang et al., 2016). However, whether basal ganglia circuits are convergently changed in ASD mouse models is an open question. Here we investigated whether loss of function of the syndromic ASD risk gene Cntnap2 alters striatal physiology and basal ganglia-dependent behaviors in mice.

Cntnap2 codes for a neurexin-like cell adhesion molecule called Contactin-associated protein-like 2 (Caspr2) (Poliak et al., 1999; Poliak et al., 2003). In mice, Caspr2 is expressed in several cortical and subcortical regions, including the striatum, from embryonic day 14 (E14) onward into adulthood (Penagarikano et al., 2011). Caspr2 is primarily localized at the juxtaparanodes of myelinated axons and is involved in the clustering of potassium channels (Poliak et al., 2003; Scott et al., 2019). In mice, in vitro work suggests that Caspr2 may also play a role in AMPAR trafficking and cell morphology (Anderson et al., 2012; Gdalyahu et al., 2015; Varea et al., 2015), and ex vivo studies indicate that it can control cell excitability and circuit synchronicity (Martin-de-Saavedra et al., 2022). Caspr2 is important during neurodevelopment and has been implicated in neuronal migration (Penagarikano et al., 2011), the maturation and function of parvalbumin-positive GABAergic interneurons (Penagarikano et al., 2011; Scott et al., 2019; Vogt et al., 2018), and the timing of myelination (Scott et al., 2019). CNTNAP2 mutations in people lead to a neurodevelopmental syndrome that can include language disorders, epilepsy, obsessive compulsive disorder, and ASD (Penagarikano & Geschwind, 2012; Rodenas-Cuadrado et al., 2014). A mouse model of this syndrome, Cntnap2-/-, has been shown to exhibit good face validity for ASD-relevant social and motor behavior alterations (Brunner et al., 2015; Dawes et al., 2018; Penagarikano et al., 2011; Scott et al., 2019).

However, the impact of Cntnap2 loss on striatal synaptic physiology and excitability and corticostriatal-dependent behaviors has not been comprehensively assessed.

The striatum is primarily composed of GABAergic striatal projection neurons (SPNs), which make up two functionally distinct output pathways: the D1-receptor expressing cells of the direct pathway (dSPNs), which project to substantia nigra pars reticulata (SNr), and the D2- receptor expressing cells of the indirect pathway (iSPNs), which project to external globus pallidus (GPe) (Calabresi et al., 2014; Gerfen & Surmeier, 2011; Kravitz et al., 2010; Tai et al., 2012). The two types of SPNs are intermixed throughout the striatum and receive excitatory glutamatergic inputs from cortex and thalamus, as well as dopaminergic input from the midbrain (Ding et al., 2008; Doig et al., 2010; Gerfen & Surmeier, 2011). Coordinated activity between the populations of SPNs in response to these inputs mediates action selection, motor learning and habit formation (Hawes et al., 2015; Santos et al., 2015; Yin et al., 2005, 2006). Although SPNs comprise upwards of 95% of the cells in striatum, there are a wide range of distinct interneurons, primarily GABAergic, that contribute significantly to the inhibitory circuitry of the striatum. Parvalbumin (PV) interneurons, which make up ∼2% of the cells in striatum, provide the largest feedforward inhibition onto SPNs (Burke et al., 2017). Changes in the number and/or function of PV interneurons have been identified in several ASD mouse models, including Cntnap2-/-mice, implicating PV circuitry as a potential common alteration across ASD mouse models (Filice et al., 2020; Juarez & Martinez Cerdeno, 2022).

To determine how loss of Cntnap2 affects striatal function, we assessed the physiology of SPNs and striatal PV-interneurons in the dorsolateral striatum and utilized a range of assays to assess striatum-associated behaviors in Cntnap2-/- mice. We find that SPNs of both the direct and indirect pathways exhibit increased corticostriatal drive, despite unchanged excitatory cortical input onto these cells. Although decreased inhibitory function has been identified in other brain regions in Cntnap2-/-mice, we find no deficit in broad or PV-specific inhibitory input onto SPNs in the case of Cntnap2 loss. Instead, we identify a significant increase in the intrinsic excitability of SPNs in Cntnap2-/- mice, in particular in dSPNs, which likely underlies the increased corticostriatal drive of these cells. Behaviorally, we find that Cntnap2-/- mice exhibit RRB-like behaviors including increased grooming, marble burying and nose poking in the holeboard assay. Cntnap2-/- mice also exhibit increased motor routine learning in the accelerating rotarod, and cognitive inflexibility in an odor-based reversal learning task. Taken together, these findings suggest that enhanced direct pathway drive may play a role in the spontaneous and learned repetitive behaviors exhibited by these mice.

Results

Cntnap2-/- SPNs exhibit increased cortical drive

Emerging evidence suggests that corticostriatal synapses are a common site of alteration in mouse models of ASD (Li & Pozzo-Miller, 2020). To test whether mice with loss of Cntnap2 exhibit changes in corticostriatal connectivity, we crossed Cntnap2-/- mice to Thy1- ChR2-YFP mice, which express channelrhodopsin in a subset of layer V pyramidal neurons (Fig. 1A) (Arenkiel et al., 2007; Poliak et al., 1999). These mice were crossed to a D1-tdTomato reporter line to visually identify dSPNs (Ade et al., 2011). We recorded from SPNs in the dorsolateral striatum (DLS), as this sensorimotor striatal subregion is implicated in the acquisition of habitual and procedural behaviors (Packard & Knowlton, 2002). Changes in physiological function in this area may be connected to the acquisition of repetitive motor behaviors in ASD (APA, 2022; Fuccillo, 2016). To simulate a train of cortical inputs, we applied ten pulses of blue light over the recording site in DLS and measured the number of action potentials (APs) fired by SPNs in the absence of synaptic blockers (Fig. 1A). We altered the light intensity to vary the probability of eliciting subthreshold depolarizations or AP firing. dSPNs were identified using tdTomato fluorescence, and tdTomato negative neurons were designated putative iSPNs.

Cntnap2-/- SPNs exhibit increased cortical drive.

(A) Left: schematic of the corticostriatal connectivity experiments. For corticostriatal excitability, cortical terminals expressing ChR2 were stimulated with 10 pulses of blue light at 10 Hz and responses were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. For corticostriatal synaptic strength, cortical terminals expressing ChR2 were stimulated with blue light at increasing intensity and synaptic currents were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. Center: 10x confocal image of the striatum from a Cntnap2+/+;D1- tdTomato;Thy1-ChR2 mouse. Right: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;D1-tdTomato;Thy1-ChR2 mouse. YFP (green) labels cell bodies and axons of a subset of layer V pyramidal neurons, tdTomato (red) labels dSPNs, and DAPI stained nuclei are in blue.

(B) Example single traces of action potentials (APs) in dSPNs evoked by cortical terminal stimulation at 20% light intensity for the indicated genotypes.

(C) Quantification (mean ± SEM) of the number of APs evoked in dSPNs at different light intensities. Cntnap2+/+ n=9 mice, 24 cells, Cntnap2-/-n=10 mice, 22 cells. Two-way ANOVA p values are shown; geno F (1, 221) = 16.75, stim F (12, 221) = 34.19, g x s F (12, 221) = 0.7682.

(D) Example single traces of action potentials (APs) in iSPNs evoked by cortical terminal stimulation at 20% light intensity for the indicated genotypes.

(E) Quantification (mean ± SEM) of the number of APs evoked in iSPNs at different light intensities. Cntnap2+/+ n=9 mice, 23 cells, Cntnap2-/-n=10 mice, 21 cells. Two-way ANOVA p values are shown; geno F (1, 234) = 8.067, stim F (12, 234) = 24.88, g x s F (12, 234) = 0.2057.

(F) Average EPSC traces from example dSPNs of each genotype induced by optogenetic cortical terminal stimulation at 14% light intensity.

(G) Quantification of EPSC amplitude evoked in dSPNs at different light intensities (line represents the mean, dots/squares represent average EPSC amplitude for each mouse). Cntnap2+/+ n=8 mice, 17 cells, Cntnap2-/- n=5 mice, 13 cells. Two-way ANOVA p values are shown; geno F (1, 33) = 0.6527, stim F (2, 33) = 11.80, g x s F (2, 33) = 0.05117.

(H) Average EPSC traces from example iSPNs of each genotype induced by optogenetic cortical terminal stimulation at 14% light intensity.

(I) Quantification of EPSC amplitude evoked in iSPNs at different light intensities (line represents mean, dots/squares represent average for mouse). Cntnap2+/+ n=6 mice, 13 cells, Cntnap2-/-n=4 mice, 11 cells. Two-way ANOVA p values are shown; geno F (1, 24) = 2.770, stim F (2, 24) = 3.466, g x s F (2, 24) = 0.1408.

We quantified the number of evoked APs at different light intensities and found that both dSPNs and iSPNs in young adult Cntnap2-/- mice exhibited significantly increased spike probability compared to wild-type (WT) SPNs (Fig. 1B-E). This suggests increased corticostriatal drive, consistent with findings in another mouse model with loss of function of the ASD-risk gene Tsc1 (Benthall et al., 2021). To test whether the enhanced spiking probability was due to a change in the synaptic excitability of Cntnap2-/-SPNs, we applied blue light of varying intensity over the recording site in DLS to evoke AMPAR-driven excitatory postsynaptic currents (EPSCs). We found that the average optically-evoked EPSC amplitude was not significantly different across a range of light intensities in Cntnap2-/-dSPNs or iSPNs compared to WT (Fig. 1F-I).

To further assess synaptic inputs, we measured the number of dendritic spines in Cntnap2-/- and WT SPNs, which are typically the sites of cortical synaptic innervation (Bouyer et al., 1984; Xu et al., 1989). To visualize spines, we injected neonate Cntnap2;Drd1a-tdTomato mice with AAV5-Syn1-GFP virus to sparsely label dSPNs and iSPNs in the DLS (Fig. S1) (Keaveney et al., 2018). We found that Cntnap2-/- SPNs in adult mice had similar spine density as WT (Fig. S1), suggesting no overall change in synapse number. Together, these results show that dSPNs and iSPNs in Cntnap2-/- mice exhibit enhanced cortically-driven spiking.

However, this is not due to a change in corticostriatal synaptic strength or overall synapse density.

Increased corticostriatal drive in Cntnap2-/- mice is not due to reduced inhibition Previous work has indicated a reduction in the number and/or function of fast-spiking parvalbumin-expressing (PV) interneurons across multiple brain regions in Cntnap2-/-mice (Ahmed et al., 2023; Antoine et al., 2019; Jurgensen & Castillo, 2015; Paterno et al., 2021; Penagarikano et al., 2011; Vogt et al., 2018). While inhibitory deficits have been identified in the cortex and hippocampus (Antoine et al., 2019; Jurgensen & Castillo, 2015), and the number of PV interneurons has been reported to be decreased in striatum (Penagarikano et al., 2011), a comprehensive assessment of inhibitory synaptic function has yet to be completed in the striatum of Cntnap2-/- mice. We first determined whether there were global deficits in inhibition onto SPNs in Cntnap2-/- mice using intrastriatal electrical stimulation to evoke inhibitory postsynaptic currents (IPSCs) (Fig. 2A). In Cntnap2-/- dSPNs, the average electrically-evoked IPSCs across a range of stimulation intensities were not different from WT dSPNs (Fig. 2B,C). However, in Cntnap2-/- iSPNs, IPSCs were, on average, slightly larger than those in WT iSPNs, although we note that the responses were variable across cells (Fig. 2D,E).

Inhibition is not reduced in Cntnap2-/- SPNs.

(A) Top: schematic of the experiment. A bipolar stimulating electrode was placed roughly 200 μm from the recording site. A range of electrical stimulation intensities was applied to the tissue while IPSCs were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. Bottom: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;D1-tdTomato mouse. tdTomato (red) labels dSPNs, DAPI stained nuclei are in blue.

(B) Average IPSC traces from example dSPNs of each genotype evoked by electrical stimulation at 1.5 (x0.01 mA) intensity for the indicated genotypes.

(C) Quantification (mean ± SEM) of IPSC amplitude in dSPNs at different stimulation intensities. Cntnap2+/+ n=17 cells from 9 mice, Cntnap2-/- n=16 cells from 9 mice. Two-way ANOVA p values are shown; geno F (1, 248) = 0.004612, stim F (7, 248) = 16.09, g x s F (7, 249) = 0.3164.

(D) Average IPSC traces from example iSPNs of each genotype evoked by electrical stimulation at 1.5 (x0.01 mA) intensity for the indicated genotypes.

(E) Quantification (mean ± SEM) of IPSC amplitude in iSPNs at different stimulation intensities. Cntnap2+/+ n=16 cells from 9 mice, Cntnap2-/- n=16 cells from 10 mice. Two-way ANOVA p values are shown; geno F (1, 240) = 11.25, stim F (7, 240) = 10.17, g x s F (7, 240) = 0.1947.

(F) Top: schematic of the experiment. PV interneuron terminals expressing ChR2 were stimulated with blue light at a range of intensities and optically-evoked IPSCs were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. Bottom: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;D1-tdTomato;PV-Cre;Ai32 mouse. YFP (green) labels PV interneurons, tdTomato (red) labels dSPNs, DAPI stained nuclei are in blue.

(G) Average IPSC traces from example dSPNs of each genotype evoked by optogenetic PV interneuron stimulation at 30% light intensity.

(H) Quantification (mean ± SEM) of IPSC amplitude in dSPNs at different light intensities. Cntnap2+/+ n=29 cells from 15 mice, Cntnap2-/- n=23 cells from 11 mice. Two-way ANOVA p values are shown; geno F (1, 504) = 0.006524, stim F (7, 504) = 14.26, g x s F (7, 504) = 0.1124.

(I) Average IPSC traces from example iSPNs of each genotype evoked by optogenetic PV interneuron stimulation at 30% light intensity.

(J) Quantification (mean ± SEM) of IPSC amplitude in iSPNs at different light intensities. Cntnap2+/+n=24 cells from 14 mice, Cntnap2-/- n=27 cells from 13 mice. Two-way ANOVA p values are shown; geno F (1, 392) = 2.638, stim F (7, 392) = 13.59, g x s F (7, 392) = 0.5088.

There are many sources of inhibition in the striatum (Burke et al., 2017), which can all be activated with electrical stimulation. To assess whether inhibition from PV interneurons specifically is altered in Cntnap2-/- mice, we crossed Cntnap2-/-;D1-tdTomato mice to Pvalb-Cre; RCL-ChR2-H134R-EYFP (Ai32) mice to express channelrhodopsin in PV interneurons (Fig. 2F) (Hippenmeyer et al., 2005; Madisen et al., 2010). We applied a blue light pulse of varying intensity over the recording site to evoke PV interneuron-specific IPSCs in SPNs, in the presence of excitatory synaptic blockers (Fig. 2F). We found that the average optically-driven IPSCs did not differ significantly in amplitude in Cntnap2-/- dSPNs or iSPNs compared to WT controls (Fig. 2G-J).

To directly measure PV neuron function, we assessed the intrinsic excitability of PV interneurons in Cntnap2-/- mice. To visualize PV interneurons for recordings, we crossed Cntnap2-/- mice to Pvalb-Cre;RCL-tdT (Ai9) mice (Fig. S2A). Plotting the number of APs fired as a function of current step size indicated that there were no significant differences in the intrinsic excitability of PV interneurons in Cntnap2-/- mice compared to controls (Fig. S2B,C). There were also no changes in intrinsic cell properties such as rheobase, membrane resistance, capacitance, resting membrane potential, or AP shape in PV interneurons in Cntnap2-/- mice (Fig. S2D-K).

Given prior reports of altered PV cell number in Cntnap2-/-mice (Paterno et al., 2021; Penagarikano et al., 2011; Vogt et al., 2018), we counted PV-expressing cells in the striatum, using immunohistochemistry and fluorescent in situ hybridization. We found no significant difference in the number of PV-positive cells in the dorsal striatum of Cntnap2-/- mice compared to WT (Fig. S3A-F and K-N). We also found no changes in the average single cell or total level of PV protein expression in the dorsal striatum (Fig. 3G-J). Overall, we did not observe significant changes in either PV interneuron number, PV expression, or PV interneuron- mediated inhibition in the adult Cntnap2-/-striatum compared to WT controls.

Intrinsic excitability is increased in Cntnap2-/- SPNs.

(A) Example AP traces in dSPNs evoked by a 200 pA current step for the indicated genotypes.

(B) Quantification (mean ± SEM) of the number of APs evoked in dSPNs at different current step amplitudes. Cntnap2+/+ n=22 cells from 8 mice, Cntnap2-/- n=23 cells from 8 mice. Two-way ANOVA p values are shown; geno F (1, 688) = 48.80, stim F (15, 688) = 107.5, g x s F (15, 688) = 1.737.

(C) Quantification (mean ± SEM) of the rheobase current in dSPNs. Dots/squares represent the rheobase current for each neuron. n is the same as in panel B. **p=0.0016, two-tailed unpaired t test.

(D-F) Quantification (mean ± SEM) of dSPN membrane resistance (D), *p=0.0328, Mann- Whitney test; membrane capacitance, p=0.2182, Mann-Whitney test; and resting membrane potential (F), p=0.9914, two-tailed unpaired t test. Dots/squares represent the average value for each neuron. n is the same as in panel B.

(G) Example AP traces in iSPNs evoked by a 200 pA current step for the indicated genotypes.

(H) Quantification (mean ± SEM) of the number of APs evoked in iSPNs at different current step amplitudes. Cntnap2+/+ n=22 cells from 8 mice, Cntnap2-/-n=21 cells from 8 mice. Two-way ANOVA p values are shown; geno F (1, 656) = 4.186, stim F (15, 656) = 118.7, g x s F (15, 656) = 0.7780.

(I) Quantification (mean ± SEM) of the rheobase current in iSPNs. Dots/squares represent the rheobase current for each neuron. n is the same as in panel H. p=0.0923, two-tailed unpaired t test.

(J-K) Quantification (mean ± SEM) of iSPN membrane resistance (J), p=0.8193, Mann-Whitney test; membrane capacitance (K), p=0.6886, two-tailed unpaired t test; and resting membrane potential (L), p=0.4859, two-tailed unpaired t test. Dots/squares represent the average value for each neuron. n is the same as in panel H.

SPN intrinsic excitability is increased in Cntnap2-/- mice

Given that the increased cortical drive onto Cntnap2-/-SPNs could not be explained by changes in excitatory or inhibitory synaptic function, we tested whether it could be due to a change in the intrinsic excitability of SPNs. To measure this, we recorded from dSPNs and iSPNs in Cntnap2-/-;D1-tdTomato mice and injected current steps of increasing amplitude. We found that Cntnap2-/- SPNs exhibited significantly increased intrinsic excitability compared to WT SPNs, which was most pronounced for dSPNs (Fig. 3A-L). Cntnap2-/- dSPNs, but not iSPNs,

also had reduced rheobase current (Fig. 3C), the minimum current required to evoke an AP, as well as increased membrane resistance (Fig. 3D). Membrane capacitance (Fig. 3E,K), resting membrane potential (Fig. 3F,L), and AP shape (Fig. S4) were not significantly changed in Cntnap2-/- SPNs. Given the lack of synaptic changes observed in Cntnap2-/- SPNs, the increase in SPN intrinsic excitability likely underlies their enhanced corticostriatal drive (see Fig. 1).

Cntnap2-/- mice exhibit increased repetitive behaviors

RRBs comprise one of the primary symptom domains of ASD (APA, 2022). Alterations in striatal circuits are thought to be involved in the manifestation of RRBs, given the striatum’s role in action selection and motor control (Estes et al., 2011; Fuccillo, 2016; Hollander et al., 2005; Langen et al., 2014). To determine whether changes in motor behavior accompanied the altered striatal physiology in Cntnap2-/- mice, we assessed locomotor activity and spontaneous repetitive behaviors using the open field, marble burying, and holeboard assays (Fig. 4A-D). In the open field, we found no significant difference in the total distance traveled, average speed, or number of rears in Cntnap2-/- mice compared to WT controls (Fig. 4E-G). We did find that Cntnap2-/- mice made significantly more entries into the center of the open field arena than WT mice, which may reflect a reduction in avoidance behavior in these mice (Fig. 4H). Manually scored grooming behavior revealed that Cntnap2-/- mice initiated more grooming bouts than WT controls in the open field (Fig. 4I).

Cntnap2-/- mice exhibit increased repetitive behaviors.

(A-D) Schematics of the behavioral assays used to measure repetitive behaviors in Cntnap2+/+ and Cntnap2-/- mice, created with BioRender.com. Male and female mice were used for all tests. (E-H) Quantification (mean ± SEM) of open field activity over 60 minutes. (E) total distance traveled, p= 0.3538, Mann-Whitney test; (F) average speed, p=0.3832, Mann-Whitney test; (G) number of rears, p=0.1892, two-tailed unpaired t test; (H) number of center entries, *p=0.0101, two-tailed unpaired t test. Cntnap2+/+ n=41 mice, Cntnap2-/-n=34 mice.

(I) Quantification (mean ± SEM) of the number of manually scored grooming bouts in the first 20 minutes of the open field test, **p=0.0034, Mann-Whitney test, Cntnap2+/+ n=41 mice, Cntnap2-/-n=34 mice.

(J) Quantification (mean ± SEM) of total marbles buried in the marble burying assay. Cntnap2+/+ n=33 mice and Cntnap2-/- n=33 mice, *p=0.0396, two-tailed unpaired t test.

(K-M) Quantification (mean ± SEM) of performance in the holeboard assay. (K) Total number of nose pokes in 10 minutes, *p=0.0212, Mann-Whitney test; (L) nose pokes in the first five minutes, p=0.4811, two-tailed unpaired t test; and (M) nose pokes in the last five minutes, *p=0.0116, two-tailed unpaired t test. Cntnap2+/+ n=25 mice, Cntnap2-/- n=22 mice.

(N) Quantification (mean ± SEM) of the frequency of movement syllables (top 25 most frequent syllables) in the open field assay defined by Keypoint-MoSeq. Cntnap2+/+n=13 mice and Cntnap2-/- n=11 mice, **p=0.0013 for grooming 6, **p=0.0013 for grooming 7, Kruskal-Wallis test with Dunn’s correction for multiple comparisons.

(O) Quantification (mean ± SEM) of the entropy of syllable transitions in the open field assay. Cntnap2+/+ n=13 mice and Cntnap2-/- n=11 mice, *p=0.0236, two-tailed unpaired t test.

(P) Accuracy of a Random Forest decoder trained on DeepLabCut basic locomotor data (Movement), Keypoint-MoSeq syllable usage data (Syllable usage), or Keypoint-MoSeq syllable transition data (Syllable transition) in distinguishing between Cntnap2+/+ and Cntnap2-/-mice. Dotted line represents chance performance.

For panels E-M and O, dots/squares represent the value for each mouse.

To further assess motor behaviors in Cntnap2-/- mice, we utilized the marble burying assay (Fig. 4C), which takes advantage of a mouse’s natural tendency to dig or bury. The number of marbles buried is used as a measure of persistent or repetitive behavior (Angoa- Perez et al., 2013). We found that Cntnap2-/- mice buried significantly more marbles on average than WT controls (Fig. 4J). Another measure of repetitive behavior, which is based on the natural exploratory behavior of mice is the holeboard assay (Fig. 4D). In this task, the number of nose pokes made into unbaited holes is recorded. We found that Cntnap2-/-mice made significantly more nose pokes within a 10-minute period than WT mice (Fig. 4K). This was largely due to increased poking during the last 5 minutes of the test (Fig. 4L-M), indicating persistent poking behavior in Cntnap2-/- mice. Together with the increased grooming identified in the open field, increased marble burying and nose poking indicate an increase in RRBs in Cntnap2-/-mice. A summary of behavior test results by genotype and sex is shown in Table S1.

To gain further insight into the spontaneous behavior profile of Cntnap2-/-mice, we utilized a combination of DeepLabCut and Keypoint-MoSeq to perform unbiased, machine learning-based assessment of general locomotion and behavior in an additional cohort of Cntnap2-/- mice (Fig. 4N-P, S5) (Mathis et al., 2018; Wiltschko et al., 2020). Again, we found that Cntnap2-/- mice did not exhibit major changes in general locomotor activity compared to WT littermates (Fig. S5A). Analysis of movement “syllables” using Keypoint-MoSeq revealed that across the 25 most frequently performed syllables, two syllables associated with grooming were performed with significantly increased frequency in Cntnap2-/- mice (Fig. 5N). Cntnap2-/-mice also had an increase in the total number of grooming bouts (Fig. S5B), replicating the findings in the manually scored cohort (see Fig. 4I). While syllable usage was generally similar between WT and Cntnap2-/- mice, transitions between syllables differed between the groups (Fig. S5C).

Cntnap2-/- mice exhibit enhanced motor learning.

(A) Schematic of the rotarod apparatus (top), and design of the task (bottom), created with BioRender.com. Mice walk to stay on the rotating rod for three 5-minute trials a day for two days at 5-40 RPM acceleration over 5 minutes, followed by three trials a day for two days at 10-80 RPM.

(B) Quantification (mean ± SEM) of accelerating rotarod performance across 12 trials for the indicated genotypes. Cntnap2+/+ n=30 mice, Cntnap2-/- n=29 mice. Two-way ANOVA p values are shown; geno F (1, 683) = 92.17, trial F (11, 683) = 58.45, g x t F (11, 683) = 2.820.

(C) Quantification (mean ± SEM) of rotarod performance on trial 1 quantified as terminal speed. Dots/squares represent the performance of individual mice. n is same as in panel B, p=0.1518, Mann-Whitney test.

(D) Quantification (mean ± SEM) of learning rate (RPM/day) calculated as the slope of the line of performance from the first trial (1) to the last trial (12) for each mouse. Dots/squares represent the learning rate for individual mice. n is same as in panel B, ***p=0.0002, two-tailed unpaired t test.

A measure of the entropy of syllable transitions revealed that Cntnap2-/-mice exhibited less entropy, suggesting less variability in the transition from one movement syllable to the next (Fig. 4O). This rigidity in motor sequence may be indicative of more restricted motor behavior overall. Finally, we tested whether a trained decoder could accurately distinguish WT and Cntnap2-/-mice using information about movement, syllable usage, or syllable transitions. The decoding models performed significantly better than chance at identifying WT and Cntnap2-/- mice based on their syllable usage and transitions, but not general locomotor activity (Fig. 4P). Together, this analysis demonstrates that while overall locomotor activity is not strongly affected in Cnantp2-/- mice, the behavior patterns of these mice are distinct from WT, reflecting enhanced presence of RRBs.

Cntnap2-/- mice exhibit enhanced motor learning

The accelerating rotarod is a striatal-dependent measure of motor coordination and learning that has been used across a range of ASD mouse models (Cording & Bateup, 2023). Changes in corticostriatal circuits have been identified in mouse models of ASD with altered performance in the task (Cording & Bateup, 2023). Given the altered corticostriatal drive in Cntnap2-/- mice, we tested whether motor coordination and learning were affected in these mice. In the rotarod test, mice learn to walk and then run to stay on a rotating rod as it increases in speed over the course of five minutes. Mice perform three trials a day for four days. In trials one through six, the rod increases in speed from five to 40 revolutions per minute (RPM), while in trials seven through 12 the rod increases from 10 to 80 RPM (Fig. 5A). Learning occurs over trials within a day, as well as across days, as the mouse develops and hones a stereotyped motor pattern to stay on the rod for increasing amounts of time (Rothwell et al., 2014; Yin et al., 2009). We found that Cntnap2-/- mice performed significantly better than WT mice in this task, in particular, in the later trials when the rod rotates at the faster 10 to 80 RPM speed (Fig. 5B).

Initial performance (terminal velocity on trial one) was not significantly different between WT and Cntnap2-/- mice (Fig. 5C), but the learning rate from trial one to trial 12 was increased in Cntnap2-/- mice (Fig. 5D). These findings expand upon previous work indicating increased performance on both steady-state and accelerating rotarod tasks utilizing slower speeds in Cntnap2-/-mice (Dawes et al., 2018; Penagarikano et al., 2011). These results also align with the increased rotarod performance seen in other ASD mouse models exhibiting enhanced corticostriatal drive (Benthall et al., 2021; Cording & Bateup, 2023).

Cntnap2-/- mice exhibit cognitive inflexibility

RRBs include not just stereotyped movements, but also insistence on sameness and perseverative interests (APA, 2022). Cognitive inflexibility, a deficit in the ability to flexibly adapt to changes in the environment and update behavior, is a manifestation of ASD and some other psychiatric disorders, which are associated with striatal dysfunction (Fuccillo, 2016). Indeed, in individuals with ASD, the severity of RRBs is associated with measures of cognitive inflexibility, and evidence from imaging studies suggests that altered corticostriatal connectivity may be present in the case of both repetitive behaviors and cognitive inflexibility (Uddin, 2021). To assess cognitive flexibility in Cntnap2-/-mice, we utilized a four-choice odor-based reversal learning assay (Johnson et al., 2016; Lin et al., 2022). Briefly, mice are trained to dig for a reward in one of four pots containing scented wood shavings (Fig. 6A). On the first day of the task (acquisition), the rewarded pot is scented with odor one (O1). Mice reach criterion when they have chosen O1 for at least eight of 10 consecutive trials. On day two, mice are given a recall test in which the rewarded pot is again scented with O1. After reaching criterion, the reversal trials begin, and the rewarded pot is scented with the previously unrewarded odor two (O2). To reach criterion, mice must learn the new association of O2 and reward and choose O2 for eight of 10 consecutive trials.

Cntnap2-/- mice exhibit cognitive inflexibility.

(A) Schematic of the four-choice odor-based reversal learning task, created with BioRender.com. Colored circles represent pots with different scented wood shavings. Yellow ring represents the food reward. Red circle in the Day 2 Reversal panel indicates a novel odor. (B-D) Quantification of parameters during acquisition learning. Mean ± SEM number of trials to reach criterion (B, at least 8 out of last 10 trials correct), p=0.5397, Mann-Whitney test; number of quadrant entries before making a choice (C), p=0.9118, Mann-Whitney test; and latency to make a choice (D), p=0.7224, two-tailed unpaired t test. Dots/squares represent the value for each mouse.

(E) Quantification (mean ± SEM) of the number of trials to reach criterion (at least 8 out of last 10 trials correct) during the recall test on day 2, p=0.3737, Mann-Whitney test.

(F-I) Quantification of parameters during reversal learning. Mean ± SEM number of trials to reach criterion (F, at least 8 out of last 10 trials correct), **p=0.0048, two-tailed unpaired t test; number of quadrant entries before making a choice (G), *p=0.0158, two-tailed unpaired t test; latency to make a choice (H), **p=0.0013, two-tailed unpaired t test; and number of trials to reach criterion after the first correct choice (I), *p=0.0183, two-tailed unpaired t test.

(J) Quantification (mean ± SEM) of the total number of errors made during reversal learning, **p=0.0034, Mann-Whitney test.

(K) Quantification (mean ± SEM) of the different error types made during reversal learning. Perseverative errors, ***p=0.0005, regressive errors **p=0.0068, novel errors, p=0.9955, irrelevant errors p=0.9988, omissions, p=>0.9999, repeated measures two-way ANOVA with Šídák’s multiple comparisons test; geno F (1, 18) = 10.91, error F (4, 72) = 53.49, g x e F (4, 72) = 4.292, animal F (18, 72) = 0.9638.

(L) Quantification (mean ± SEM) of the percent of choices made for each odor, binned across three trials, during reversal learning. Odor 1 was rewarded during acquisition learning and odor 2 was rewarded during reversal learning.

For panels B-L, n=10 Cntnap2+/+ mice and 10 Cntnap2-/- mice.

During acquisition, Cntnap2-/- mice performed similarly to WT controls, not differing in the average number of trials needed to reach criterion, the number of quadrant entries made before making a choice, or the latency to choose a pot (Fig. 6B-D). On day two, Cntnap2-/- mice performed similarly to controls during recall, demonstrating successful consolidation of the odor- reward pairing (Fig. 6E). However, we found that Cntnap2-/- mice exhibited a deficit in reversal learning, requiring significantly more trials on average than WTs to reach criterion once the odor-reward pairing was changed (Fig. 6F). Interestingly, during reversal, Cntnap2-/- mice made fewer quadrant entries before making a digging choice and exhibited significantly decreased latency to make a choice compared to controls (Fig. 6G,H). Even after the first correct choice of O2 during reversal, Cntnap2-/- mice took more trials to reach criterion than WTs (Fig. 6I). In terms of errors, Cntnap2-/- mice made more reversal errors than WT mice (Fig. 6J), in particular perseverative (continuing to choose O1) and regressive (choosing O1 after correctly choosing O2 once) errors (Fig. 6K). Cntnap2-/-mice did not differ from WT controls in choices of the novel (newly introduced during reversal) or irrelevant (never rewarded) odors, or in the number of omitted trials (timing out without making a choice) (Fig. 6K). Instead, the persistence in choosing O1, even after at least one correct choice of O2, drove the cognitive inflexibility in these mice (Fig. 6L). This persistence in choice may be reflective of the broader scope of RRBs in Cntnap2-/- mice.

Discussion

In this study we tested whether loss of the ASD risk gene Cntnap2 results in altered striatal physiology or striatal-dependent behavior. We found that both direct and indirect pathway SPNs exhibited significantly increased cortical drive in Cntnap2-/- mice. This change was likely not due to a change in excitatory or inhibitory synapses, as excitatory cortical input onto SPNs was unchanged and there were no significant deficits in inhibition onto SPNs in these mice. Instead, loss of Cntnap2 resulted in a significant increase in the intrinsic excitability of SPNs, in particular dSPNs. At the behavioral level, Cntnap2-/- mice exhibited repetitive behaviors including increased grooming, nose poking, and marble burying. These mice also had enhanced motor learning, performing significantly better than controls in the accelerating rotarod task. Finally, Cntnap2-/- mice exhibited cognitive inflexibility in the four-choice reversal learning task.

Cellular phenotypes of Cntnap2 loss

The loss of Caspr2 has a variable impact on intrinsic excitability across brain regions and cell types. The increased intrinsic excitability that we identified in SPNs has also been observed in Purkinje cells of the cerebellum (Fernandez et al., 2021) and pyramidal cells of the cortex (Antoine et al., 2019; Cifuentes-Diaz et al., 2023). However, we note hypoactivity (Brumback et al., 2018) or unchanged excitability (Lazaro et al., 2019) of pyramidal cells in some cortical regions in Cntnap2-/- mice. Caspr2 is involved in the clustering of voltage-gated potassium channels, in particular at the juxtaparanodes of myelinated axons (Poliak et al., 1999; Poliak et al., 2003) and the axon initial segment (Inda et al., 2006). Indeed, there are profound deficits in the clustering of Kv1-family channels in Cntnap2-/- mice, particularly Kv1.2 channels (Scott et al., 2019). These channels play an important role in regulating the intrinsic excitability of SPNs (Nisenbaum et al., 1994), and when blocked, result in decreased rheobase and increased firing (Shen et al., 2004). The loss of Caspr2 in Cntnap2-/- mice may result in improper localization of these channels, which could contribute to the decreased rheobase and increased SPN excitability we observed. While Caspr2 has most often been discussed for this interaction with Kv1-family channels, recent work has shown that it can also play a role in calcium signaling. The extracellular domain of Caspr2, after undergoing cleavage through proteolytic processing, can interact with the calcium extrusion pump PMCA2, promoting calcium export and likely decreasing cell excitability (Martin-de-Saavedra et al., 2022). This process, which increases with neuronal activity, may be disrupted in Cntnap2-/- mice, which could also contribute to the increased intrinsic excitability that we found in SPNs.

Prior studies of Cntnap2-/- mice have identified changes in the number of PV-expressing interneurons in the cortex (Penagarikano et al., 2011; Vogt et al., 2018), hippocampus (Paterno et al., 2021; Penagarikano et al., 2011) and striatum (Penagarikano et al., 2011). However, this finding is inconsistent across studies, as others have reported no change in the number of PV interneurons in these regions (Ahmed et al., 2023; Lauber et al., 2018; Scott et al., 2019). One possible explanation for this disparity is altered PV protein expression in Cntnap2-/- mice such that immunoreactivity varies in cell counting assessments. This is supported by the finding that the number of Vicia Villosa Agglutinin-positive (VVA+) perineuronal nets that preferentially surround PV cells is unchanged in Cntnap2-/- mice, even when PV immunoreactivity varies (Hartig et al., 1992; Haunso et al., 2000; Lauber et al., 2018). Parvalbumin, a Ca2+ buffer, plays an important role in the intrinsic fast-spiking properties of PV interneurons, such that a reduction in PV protein expression is known to change PV intrinsic function (Orduz et al., 2013). However, altered intrinsic properties of PV interneurons has also been variably reported across brain regions and studies of Cntnap2-/- mice, with subtle changes in PV firing properties reported in the developing striatum (Ahmed et al., 2023) and adult cortex (Vogt et al., 2018), but unchanged in the hippocampus (Paterno et al., 2021) and medial prefrontal cortex (Lazaro et al., 2019). In this study, we find no significant change in the number of PV interneurons or the striatal expression of PV protein in Cntnap2-/-mice. Following from this, we find no significant deficit in inhibition onto SPNs nor do we find any differences in PV interneuron intrinsic excitability in these mice. Interestingly, we find that broad inhibition measured through intrastriatal stimulation is enhanced specifically onto iSPNs in Cntnap2-/- mice. This change is likely driven by altered function or connectivity of another type of striatal interneuron, or other SPNs, which provide lateral inhibition onto iSPNs (Burke et al., 2017; Koos et al., 2004).

Loss of Cntnap2 alters striatal-dependent behaviors

The striatum can be separated into functionally distinct subregions. We focused on the dorsal striatum in this study because of its role in controlling motor and cognitive functions (Voorn et al., 2004), which are relevant to ASD (Fuccillo, 2016; Subramanian et al., 2017). The dorsal striatum can be further subdivided into the dorsomedial striatum (DMS) and the dorsolateral striatum (DLS), with the former considered an associative region involved in goal- directed action-outcome learning and the latter implicated in the acquisition of habitual or procedural behaviors (Packard & Knowlton, 2002). We focused on cell function in DLS as stereotyped, perseverative or persistent behaviors likely recruit DLS circuitry (Evans et al., 2024; Fuccillo, 2016). In the accelerating rotarod assay, learning and performance in the task has been associated with changes in the DLS. Positive modulation of the firing rate of DLS neurons occurs during rotarod training, in particular in later trials of the task, and synaptic potentiation of DLS SPNs in late training is necessary for intact performance (Yin et al., 2009). In line with this, lesions of the DLS impair both early and late rotarod learning (Yin et al., 2009). We found that Cntnap2-/- mice had increased rotarod performance, most notably at the later stages when DLS function is strongly implicated. Functionally, we also found increased cortical drive of DLS SPNs in these mice, in particular of dSPNs, a change that was associated with increased rotarod performance in another mouse model with disruption of an ASD-risk gene (Benthall et al., 2021). Together, this supports a connection between the functional change observed in DLS SPN physiology and the increased motor routine learning in Cntnap2-/- mice.

In terms of restricted, repetitive behaviors, we replicated prior studies showing increased spontaneous grooming in Cntnap2-/- mice (Penagarikano et al., 2011). Early evidence implicates the striatum in the control of the syntax or sequence of movements in a rodent grooming bout, such that very small lesions of DLS are capable of disrupting grooming (Cromwell & Berridge, 1996). However, recent work has also outlined roles for cellular modulation in DMS and ventral striatal Islands of Calleja in the control of grooming behavior (Ramirez-Armenta et al., 2022; Zhang et al., 2021). Cntnap2-/- mice also exhibited increased marble burying and nose poking. The precise neurobiological substrates of these behaviors are yet unclear, but evidence linking increases in these behaviors to changes in cortico-striatal and amygdala-striatal function supports the notion that these behaviors may fit into a broader basal ganglia-associated RRB- like domain (Albelda & Joel, 2012; Lee et al., 2024).

In the four-choice reversal learning task, Cntnap2-/- mice showed no differences during the acquisition phase, suggesting that there were no broad deficits in learning. However, in the reversal stage of the task, Cntnap2-/- mice took significantly more trials to learn a new odor- reward pairing, owing primarily to continued choice of the previously rewarded odor. The DMS and ventral striatum (nucleus accumbens) have been shown to play an important role in reversal learning (Izquierdo et al., 2017), and in the four choice task specifically (Delevich et al., 2022). Additionally, decreased dopamine release in the DLS is associated with deficits in reversal learning in this task (Kosillo et al., 2019; Lin et al., 2022). Together, the learning phenotypes seen in Cntnap2-/- mice in the accelerating rotarod and reversal learning assay share an underlying rigidity in behavioral choice. In both cases, changes in striatal circuits likely underlie the repetitive, stereotyped behaviors.

In summary, our results fit into a model whereby divergent cellular changes in the striatum driven by a functionally diverse set of ASD risk genes similarly enhance corticostriatal drive, in particular, of the direct pathway. This in turn may facilitate striatal-dependent motor routine learning and behavioral perseveration. We speculate that a shared gain-of-function in striatal circuits may play a role in the formation of perseverative or repetitive behaviors in a sub- set of ASDs more broadly.

Experimental procedures

Mice

All animal procedures were conducted in accordance with protocols approved by the University of California, Berkeley Institutional Animal Care and Use Committee (IACUC) and Office of Laboratory Animal Care (OLAC) (AUP-2016-04-8684-2). Cntnap2-/- mice and littermate Cntnap2+/+controls with the following alleles were utilized for each experiment.

Mice were group housed on a 12 h light/dark cycle (dark cycle 9:00 AM – 9:00 PM) and given ad libitum access to standard rodent chow and water. Both male and female animals were used for experimentation. The ages, sexes, and numbers of mice used for each experiment are indicated in the respective method details and figure legends. All mice used for experiments were heterozygous or hemizygous for the Drd1a-tdTomato, Thy1-ChR2-YFP, PV-Cre, Ai32, or Ai9 transgenes to avoid potential physiological or behavioral alterations.

Electrophysiology

Mice (P50-60) were briefly anesthetized with isoflurane and perfused transcardially with ice-cold ACSF (pH = 7.4) containing (in mM): 127 NaCl, 25 NaHCO3, 1.25 NaH2PO4, 2.5 KCl, 1 MgCl2, 2 CaCl2, and 25 glucose, bubbled continuously with carbogen (95% O2 and 5% CO2). Brains were rapidly removed and coronal slices (275 μm) were cut on a VT1000S vibratome (Leica) in oxygenated ice-cold choline-based external solution (pH = 7.8) containing (in mM): 110 choline chloride, 25 NaHCO3, 1.25 NaHPO4, 2.5 KCl, 7 MgCl2, 0.5 CaCl2, 25 glucose, 11.6 sodium ascorbate, and 3.1 sodium pyruvate. Slices were recovered in ACSF at 36°C for 15 min and then kept at room temperature (RT) before recording. Recordings were made with a MultiClamp 700B amplifier (Molecular Devices) at RT using 3-5 MOhm glass patch electrodes (Sutter, #BF150-86-7.5). Data were acquired using ScanImage software, written and maintained by Dr. Bernardo Sabatini (https://github.com/bernardosabatini/SabalabAcq). Traces were analyzed in Igor Pro (Wavemetrics). Recordings with a series resistance > 25 MOhms or holding current above 200 pA were rejected.

Current-clamp recordings

Current clamp recordings were made using a potassium-based internal solution (pH = 7.4) containing (in mM): 135 KMeSO4, 5 KCl, 5 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 10 phosphocreatine, and 1 EGTA. For corticostriatal excitability experiments, optogenetic stimulation consisted of a full-field pulse of blue light (470 nm, 0.5 ms pulse width, CoolLED) through a 63x objective (Olympus, LUMPLFLN60XW). Light power was linear over the range of intensities tested. No synaptic blockers were included. For intrinsic excitability experiments (both SPN and PV interneuron), NBQX (10 μM, Tocris, #1044), CPP (10 μM, Tocris, #0247) and picrotoxin (50 μM, Abcam, #120315) were added to the external solution to block synaptic transmission. One second depolarizing current steps were applied to induce action potentials.

No holding current was applied to the membrane.

Voltage-clamp recordings

Voltage-clamp recordings were made using a cesium-based internal solution (pH = 7.4) containing (in mM): 120 CsMeSO4, 15 CsCl, 10 TEA-Cl, 8 NaCl, 10 HEPES, 1 EGTA, 5 QX-314, 4 Mg-ATP, and 0.3 Na-GTP. Recordings were acquired with the amplifier Bessel filter set at 3 kHz. Corticostriatal synaptic stimulation experiments to measure evoked AMPA-mediated EPSCs were performed in picrotoxin (50 μM), and optogenetic stimulation consisted of a full- field pulse of blue light (470 nm, 0.15 ms pulse width) through a 63x objective. Synaptic stimulation experiments to measure evoked IPSCs were performed in NBQX (10 μM) and CPP (10 μM). For electrically evoked IPSCs, a concentric bipolar stimulating electrode (FHC, #30214) was placed in dorsal striatum, roughly 200 μm medial to the recording site in dorsolateral striatum, and a 0.15 ms stimulus was applied. For PV-interneuron optically evoked IPSCs, a full-field pulse of blue light (470 nm, 0.15 ms pulse width) was applied through a 63x objective at the recording site.

Dendritic imaging and spine analysis

Neonatal (P1-3) Cntnap2-/-;D1-tdT and Cntnap2+/+;D1-tdT mice were cryoanesthetized and injected bilaterally with 200 nL AAV1.hSyn.eGFP.WPRE.bGH (Penn Vector Core, #p1696 (Keaveney et al., 2018)), diluted 1:75 in saline to achieve sparse transduction. Injections were targeted to the dorsal striatum, with coordinates approximately 1.3 mm lateral to midline, 2.0 mm posterior to bregma, and 1.5 mm ventral to the head surface. At P50-60, mice were anesthetized with isoflurane and transcardial perfusion was performed with 10 mL of 1x PBS followed by 10 mL of ice cold 4% PFA (EMS, #15710-S) in 1x PBS. Brains were post-fixed in 4% PFA in 1x PBS overnight at 4° C. 80 μm coronal sections were made using a freezing microtome (American Optical, AO 860) and stored in 1x PBS at 4° C. Sections were blocked for 1 hour at RT in BlockAid (ThermoFisher, #B10710) and incubated for 48 hours with gentle shaking at 4° C with antibodies against GFP (1:2500, Abcam, #13970) and RFP (1:1000, Rockland (VWR, #600-401-379) diluted in PBS-Tx (1x PBS with 0.25% Triton X-100 (Sigma, #T8787). Sections were washed 3 x 10 min in PBS-Tx and incubated with gentle shaking for 1 hour at RT with Alexa Fluor 488 and 546 secondary antibodies (1:500, Invitrogen, #A11039, #A11035). Sections were washed 3 x 10 min in 1x PBS and mounted onto SuperFrost slides (VWR, #48311- 703) using VECTASHIELD HardSet Antifade Mounting Medium (Vector Laboratories, #H-1400-10). Z stack images of individual dendrites were taken on a confocal microscope (Olympus FLUOVIEW FV3000) with a 60x oil immersion objective (Olympus #1- U2B832) at 2.5x zoom with a step size of 0.4 μm and deconvoluted using Olympus CellSens software. To quantify spine density, dendrites and spines were reconstructed using the FilamentTracer module in Imaris software (Oxford Instruments). The spine density of each dendrite was calculated using Imaris. Dendrites analyzed varied in total length, but excluded the most proximal and distal portions of the dendrite.

Brain sectioning and immunohistochemistry

Adult mice were perfused as above and brains were post-fixed with 4% paraformaldehyde overnight, then sectioned coronally at 30 μm. For immunohistochemistry,

individual wells of sections were washed for 3 x 5 min with 1x PBS, then blocked for 1 hour at RT with BlockAid blocking solution. Primary antibodies diluted in PBS-Tx were added and tissue was incubated for 48 hours with gentle shaking at 4° C. Sections were then washed 3 x 10 min with PBS-Tx. Secondary antibodies diluted 1:500 in PBS-Tx were added and incubated with gentle shaking for 1 hour at RT. Sections were washed 3 x 10 min in 1x PBS. Sections were mounted onto SuperFrost slides (VWR, #48311- 703) and coverslipped with VECTASHIELD HardSet with DAPI (Vector Laboratories, #H-1500-10) or VECTASHIELD HardSet Antifade Mounting Medium (Vector Laboratories, #H-1400-10). The following antibodies were used: mouse anti-PV (1:1000, Sigma, #P3088), rabbit anti-PV (1:1000, Abcam, #11427), anti-RFP (1:500, Rockland, #600-401-379), Alexa Fluor 405, 488 and 546 conjugated secondary antibodies (1:500, Invitrogen, #A-31553, #A-11001, #A-11003, and #A-11035).

PV cell counting

To count PV+ interneurons, Z-stack images of immunostained striatal sections were taken on a confocal microscope (Olympus FLUOVIEW FV3000) with a 10x or 20x objective (Olympus # 1-U2B824 or Olympus # 1-U2B825) and step size of 2 μm. For quantification, image stacks were Z-projected to maximum intensity using Fiji (ImageJ) and cropped to a 400 μm x 400 μm image in anatomically matched sections of the DLS. All PV-expressing cells within this region were counted using the ROI manager tool in ImageJ. Designation of a cell as PV positive was determined by the experimenter and consistently maintained across animals. Experimenter was blind to genotype and ROIs were made on the DAPI channel to avoid selecting regions based on PV expression. To quantify bulk PV fluorescence, ROIs were manually defined in ImageJ using the Freehand tool to cover as much of the DLS as possible, and mean fluorescence intensity was measured. To quantify individual cell PV fluorescence, ROIs were manually defined around every PV positive cell in the previosuly drawn DLS ROI using the Freehand tool, and mean fluorescence intensity was measured.

Western Blot

Adult mice (P48-55) were deeply anesthetized with isoflurane and decapitated. Brains were rapidly dissected and 1.5 mm dorsal striatum punches (Biopunch, Ted Pella, #15111-15) were collected from both hemispheres, flash-frozen in liquid nitrogen, and stored at −80° C. On the day of analysis, frozen samples were sonicated (QSonica Q55) until homogenized in 200 μl lysis buffer containing 1% SDS in 1x PBS with Halt phosphatase inhibitor cocktail (Thermo Fisher Scientific, #PI78420) and Complete mini EDTA-free protease inhibitor cocktail (Roche, #4693159001). Sample homogenates were then boiled on a heat block at 95° C for 5 min and allowed to cool to RT. Total protein content was determined using a BCA assay (Thermo Fisher Scientific, #23227). Following the BCA assay, protein homogenates were mixed with 4x Laemmli sample buffer (BioRad, #161-0747). 12.5μg of total protein per sample were then loaded onto 12% Criterion TGX gels (BioRad, #5671044) and run at 65 V. Proteins were transferred to a PVDF membrane (BioRad, #1620177) at 11 V for 14 hours at 4° C using the BioRad Criterion Blotter (BioRad, #1704070). Membranes (BioRad, #1620177) were briefly reactivated in methanol and rinsed in water 3x. After rinsing, membranes were blocked in 5% milk in 1x TBS with 1% Tween (TBS-Tween) for 1 hour at RT before being incubated with primary antibodies diluted in 5% milk in TBS-Tween overnight at 4° C. The following day, after 3 x 10 min washes with TBS-Tween, membranes were incubated with secondary antibodies for 1 hour at RT. Following 6 × 10 min washes, membranes were incubated with chemiluminescence substrate (PerkinElmer #NEL105001EA) for 1 min and exposed to Amersham Hyperfilm ECL (VWR, #95017-661).

Bands were quantified by densitometry using ImageJ software. GAPDH was used to normalize protein content and data is expressed as a percentage of control within a given experiment. The following antibodies were used: anti-Caspr2 (1:5000, Abcam, #153856), anti- PV (1:2500, Abcam, #11427), anti-GAPDH (1:5000, Cell Signaling, #51745S), and anti-rabbit goat HRP conjugate (1:5000, BioRad, #1705046).

In situ hybridization

Fluorescent in situ hybridization was performed to quantify Pvalb mRNA expression in the striatum of Cntnap2+/+ and Cntnap2-/-mice. Mice were briefly anesthetized with isoflurane and brains were harvested, flash-frozen in OCT mounting medium (Thermo Fisher Scientific, #23-730-571) on dry ice and stored at -80° C for up to 6 months. 16 µm sections were collected using a cryostat (Thermo Fisher Scientific, Microm HM 550), mounted directly onto Superfrost Plus glass slides (VWR, #48311-703) and stored at -80° C for up to 6 months. In situ hybridization was performed according to the protocols provided with the RNAscope Multiplex Fluorescent Reagent Kit (ACD, #323100). Drd1a mRNA was visualized with a probe in channel 2 (ACD, #406491-C2) and Pvalb mRNA in channel 3 (ACD, #421931-C3). After incubation, sections were secured on slides using ProLong Gold Antifade Mountant with DAPI (Invitrogen, P36935) and 60 x 24 mm rectangular glass coverslips (VWR, #16004-096). Sections were imaged on an Olympus FluoView 3000 confocal microscope using a 10x objective with 1.5x zoom and a step size of 2 µm. Pvalb-expressing cells were quantified across the entire striatum using the ROI manager tool in ImageJ. A cell was considered Pvalb positive if over 50% of the cell contained fluorescent puncta when compared to the DAPI channel. Experimenter was blind to genotype.

Behavioral analysis

All behavior studies were carried out in the dark phase of the light cycle under red lights (open field) or white lights (marble burying, holeboard, rotarod, and four choice reversal learning). Mice were habituated to the behavior testing room for at least 30 min prior to testing. Mice were given at least one day between different tests. All behavior equipment was cleaned between each trial and mouse with 70% ethanol and rinsed in diluted soap followed by water at the end of the day. If male and female mice were to be tested on the same day, male mice were run first then returned to the housing room, after which all equipment was thoroughly cleaned prior to bringing in female mice for habituation. Behavioral tests were performed with young adult male and female mice (7-11 weeks old). The experimenter was blind to genotype throughout the testing and scoring procedures.

Open field assay

Exploratory behavior in a novel environment and general locomotor activity were assessed by a 60 min session in an open field chamber (40 cm L x 40 cm W x 34 cm H) made of transparent plexiglass. Horizontal infrared photobeams (Stoelting, 60001-02A) were positioned to detect rearing. The mouse was placed in the bottom right-hand corner of the arena and behavior was recorded using an overhead camera and analyzed using ANY-maze software (Stoelting). An observer manually scored self-grooming behavior during the first 20 minutes of the test. A grooming bout was defined as an unbroken series of grooming movements, including licking of body, paws, or tail, as well as licking of forepaws followed by rubbing of face with paws.

Open field assay with DeepLabCut Keypoint-MoSeq analysis

Mice were placed in the open field arena and video recorded with a monochrome camera (FLIR Grasshopper 3, GS3-U3-41C6NIR-C) and a 16 mm wide angle lens (Kowa, LM16HC) placed above the arena from a height of 50 cm. To extract the body part (keypoint) coordinates from the video recordings, DeepLabCut (DLC) 2.3.4 (Mathis, et al. 2018; Nath, et al. 2019) was used. Fourteen body parts including nose, head, left ear, right ear, left forelimb, right forelimb, spine 1, spine 2, spine 3, left hindlimb, right hindlimb, tail 1, tail 2, and tail 3 were manually labeled on a small subset of the video frames. A DLC model was then trained using the annotated frames to label those 14 body parts for all videos recorded. The total distance traveled, and number of center entries were calculated using the coordinate of bodypart tail 1.

Discrete behavior syllables were extracted using Keypoint-MoSeq 0.4.4 (Weinreb, et al. 2023). Syllable usage and transition data were obtained using built-in functions of the Keypoint-MoSeq package. Decoding and entropy analysis were performed using customized Python 3.9 script. Code available upon request in Github. Entropy was calculated using the following equation, where ui denotes the frequency of the syllable i and pi,j denotes the transition probability from syllable i to syllable j.:

Marble burying assay

The marble burying assay was used to test for repetitive behavior. 20 black marbles were arranged in an orderly 4 x 5 grid on top of 5 cm of clean corn cob bedding in a standard mouse cage. Overhead room lights were on and white noise was played to induce mild stress. Mice were placed in the cage with the marbles for 30 minutes. The number of unburied marbles (>50% exposed) was recorded after the session.

Holeboard assay

The holeboard assay was used to measure exploratory and repetitive behavior. The holeboard apparatus consisted of a smooth, flat, opaque gray plastic platform, suspended 10 cm from the base by four plastic pegs in each corner. The board contained 16 evenly spaced 2 cm diameter holes and was surrounded by a 30 cm high clear plastic square encasing. During testing, mice were placed into the center of the holeboard. Mice explored the board for 10 minutes while video was recorded from both an above and side-view camera. Videos were used post-hoc to manually count and map the number of nose pokes made during the task. Nose pokes were defined as the mouse’s nose passing through the board barrier when viewed through the side-view camera.

Accelerating rotarod assay

The accelerating rotarod test was used to examine motor coordination and learning.

Mice were trained on a rotarod apparatus (Ugo Basile, #47650) for four consecutive days. Three trials were completed per day with a 5 min break between trials. The rotarod was accelerated from 5-40 revolutions per minute (rpm) over 300 s for trials 1-6 (days 1 and 2), and from 10-80 rpm over 300 s for trials 7-12 (days 3 and 4). On the first testing day, mice were first acclimated to the apparatus by being placed on the rotarod rotating at a constant 5 rpm for 60 s and returned to their home cage for 5 min prior to starting trial 1. Latency to fall, or to rotate off the top of the rotarod barrel, was measured by the rotarod stop-trigger timer.

Four choice odor-based reversal learning test

The four-choice odor-based reversal learning test was used to assess learning and cognitive flexibility. Animals were food restricted for 6 days in total, with unrestricted access to drinking water, and maintained at 90-95% of ad lib feeding body weight. Food was given at the end of the day once testing was completed. Food restriction and introduction to Froot Loop cereal piece (Kellogg’s, Battle Creek, MI) began 48 hours before pre-training. The four-choice test was performed in a custom-made square box (30.5 cm L × 30.5 cm W × 23 cm H) constructed of clear acrylic. Four internal walls 7.6 cm wide partially divided the arena into four quadrants. A 15.2 cm diameter removable cylinder fit in the center of the maze and was lowered between trials (after a digging response) to isolate the mouse from the rest of the maze. Odor stimuli were presented mixed with wood shavings in white ceramic pots measuring 7.3 cm in diameter and 4.5 cm deep. All pots were sham baited with a piece of Froot Loop cereal secured underneath a mesh screen at the bottom of the pot. This was to prevent mice from using the odor of the Froot Loop to guide their choice. The apparatus was cleaned with 2.5% acetic acid followed by water and the pots were cleaned with 70% ethanol followed by water between mice. The apparatus was cleaned with diluted soap and water at the end of each testing day.

On the first habituation day of pre-training (day 1), animals were allowed to freely explore the testing arena for 30 min and consume small pieces of Froot Loops placed inside empty pots positioned in each of the four corners. On the second shaping day of pre-training (day 2), mice learned to dig to find cereal pieces buried in unscented coarse pine wood shavings (Harts Mountain Corporation, Secaucus, NJ). A single pot was used and increasing amounts of unscented wood shavings were used to cover each subsequent cereal reward. The quadrant containing the pot was alternated on each trial and all quadrants were rewarded equally. Trials were untimed and consisted of (in order): two trials with no shavings, two trials with a dusting of shavings, two trials with the pot a quarter full, two trials with the pot half full, and four trials with the cereal piece completely buried by shavings. The mouse was manually returned to the center cylinder between trials.

On the days for odor discrimination (day 3, acquisition) and reversal (day 4), wood shavings were freshly scented on the day of testing. Anise extract (McCormick, Hunt Valley, MD) was used undiluted at 0.02 ml/g of shavings. Clove, litsea, and eucalyptus oils (San Francisco Massage Supply Co., San Francisco, CA) were diluted 1:10 in mineral oil and mixed at 0.02 ml/g of shavings. Thymol (thyme; Alfa Aesar, A14563) was diluted 1:20 in 50% ethanol and mixed at 0.01 ml/g of shavings. During the discrimination phase (day 3), mice had to discriminate between four pots with four different odors and learn which one contained a buried food reward. Each trial began with the mouse confined to the start cylinder. Once the cylinder was lifted, timing began, and the mouse could freely explore the arena until it chose to dig in a pot. Digging was defined as purposefully moving the shavings with both front paws. A trial was terminated if no choice was made within 3 min and recorded as omission. Criterion was met when the animal completed eight out of ten consecutive trials correctly. The spatial location of the odors was shuffled on each trial. The rewarded odor during acquisition was anise.

The first four odor choices made during acquisition were analyzed to determine innate odor preference by the percentage of choices for a given odor: Cntnap2+/+ mice: 60% thyme, 25% anise, 12.5% clove, and 2.5% litsea. Cntnap2-/-mice: 47.5% thyme, 45% anise, 7.5% clove, 0% litsea. We note that both Cntnap2+/+ and Cntnap2-/-mice exhibited the strongest innate preference for thyme, an unrewarded odor. There were no significant differences in innate odor preference.

The reversal phase of the task was carried out on day 4. Mice first performed the task with the same rewarded odor as the discrimination day to ensure they learned and remembered the task. After reaching criterion on recall (eight out of ten consecutive trials correct), the rewarded odor was switched, and mice underwent a reversal learning test in which a previously unrewarded odor (clove) was rewarded. A novel odor (eucalyptus) was also introduced, which replaced thyme. Perseverative errors were choices to dig in the previously rewarded odor that was no longer rewarded. Regressive errors were choosing the previously rewarded odor after the first correct choice of the newly rewarded odor. Novel errors were choices to dig in the pot with the newly introduced odor (eucalyptus). Irrelevant errors were choices to dig in the pot that had never been rewarded (litsea). Omissions were trials in which the mouse failed to make a digging choice within 3 min from the start of the trial. Total errors were the sum of perseverative, regressive, irrelevant, novel, and omission errors. Criterion was met when the mouse completed eight out of ten consecutive trials correctly. The spatial location of the odors was shuffled on each trial.

Quantification and statistical analysis

Experiments were designed to compare the main effect of genotype. The sample sizes were based on prior studies and are indicated in the figure legend for each experiment.

Whenever possible, quantification and analyses were performed blind to genotype. GraphPad Prism version 10 was used to perform statistical analyses. The statistical tests and outcomes for each experiment are indicated in the respective figure legend. Two-tailed unpaired t tests were used for comparisons between two groups. For data that did not pass the D’Agostino & Pearson normality test, a Mann-Whitney test was used. Two-way ANOVAs were used to compare differences between groups for experiments with two independent variables. Statistical significance was defined in the figure panels as follows: *p < 0.05, **p < 0.01, ***p < 0.001.

eLife Assessment

This valuable and well-executed study describes how deletion of the autism spectrum disorder risk gene CNTNAP2 in mice increases dorsolateral striatal projection neuron excitability and promotes repetitive behaviors and cognitive inflexibility. The evidence supporting this claim is solid, although additional experimental evidence would strengthen claims of how corticostriatal activity is altered and linked to behavioral changes. The study provides a potential cellular explanation for the repetitive and inflexible behavior in Cntnap2 knockout mice and CNTNAP2 disorder in humans, which would interest both basic and translational neuroscientists.

Reviewer #1 (Public Review):

Summary:

Cording et al. investigated how deletion of CNTNAP2, a gene associated with autism spectrum disorder, alters corticostriatal engagement and behavior. Specifically, the authors present slice electrophysiology data showing that striatal projection neurons (SPNs) are more readily driven to fire action potentials in response to stimulation of corticostriatal afferents, and this is due to increases in SPN intrinsic excitability rather than changes in excitatory or inhibitory synaptic inputs. The authors show that CNTNAP2 mice display repetitive behaviors, enhanced motor learning, and cognitive inflexibility. Overall the authors' conclusions are supported by their data, but a few claims could use some more evidence to be convincing.

Strengths:

The use of multiple behavioral techniques, both traditional and cutting-edge machine learning-based analyses, provides a powerful means of assessing repetitive behaviors and behavioral transitions/rigidity. Characterization of both excitatory and inhibitory synaptic responses in slice electrophysiology experiments offers a broad survey of the synaptic alterations that may lead to increased corticostriatal engagement of SPNs.

Weaknesses:

(1) The authors conclude that increased cortical engagement of SPNs is due to changes in SPN intrinsic excitability rather than synaptic strength (either excitatory or inhibitory). One weakness is that only AMPA receptor-mediated responses were measured. Though the holding potential used for experiments in Figure 1F-I wasn't clear, recordings were presumably performed at a hyperpolarized potential that limits NMDA receptor-mediated responses. Because the input-output experiments used to conclude that corticostriatal engagement of SPNs is elevated (Figure 1B-E) were conducted in the current clamp, it is possible that enhanced NMDA receptor engagement contributed to increased SPN responses to cortical stimulation. Confirming that NMDA receptor-mediated EPSC components are not altered would strengthen the main conclusion.

(2) Data clearly show that SPN intrinsic excitability is increased in knockout mice. Given that CNTNAP2 has been linked to potassium channel regulation, it would be helpful to show and quantify additional related electrophysiology data such as negative IV curve responses and action potential hyperpolarization.

(3) As it stands, the reported changes in dorsolateral striatum SPN excitability are only correlative with reported changes in repetitive behaviors, motor learning, and cognitive flexibility.

Reviewer #2 (Public Review):

Summary:

This is an important study characterizing striatal dysfunction and behavioral deficits in Cntnap2-/- mice. There is growing evidence suggesting that striatal dysfunction underlies core symptoms of ASD but the specific cellular and circuit level abnormalities disrupted by different risk genes remain unclear. This study addresses how the deletion of Cntnap2 affects the intrinsic properties and synaptic connectivity of striatal spiny projection neurons (SPN) of the direct (dSPN) and indirect (iSPN) pathways. Using Thy1-ChR2 mice and optogenetics the authors found increased firing of both types of SPNs in response to cortical afferent stimulation. However, there was no significant difference in the amplitude of optically-evoked excitatory postsynaptic currents (EPSCs) or spine density between Cntnap2-/- and WT SPNs, suggesting that the increased corticostriatal coupling might be due to changes in intrinsic excitability. Indeed, the authors found Cntnap2-/- SPNs, particularly dSPNs, exhibited higher intrinsic excitability, reduced rheobase current, and increased membrane resistance compared to WT SPNs. The enhanced spiking probability in Cntnap2-/- SPNs is not due to reduced inhibition. Despite previous reports of decreased parvalbumin-expressing (PV) interneurons in various brain regions of Cntnap2-/- mice, the number and function (IPSC amplitude and intrinsic excitability) of these interneurons in the striatum were comparable to WT controls.

This study also includes a comprehensive behavioral analysis of striatal-related behaviors. Cntnap2-/- mice demonstrated increased repetitive behaviors (RRBs), including more grooming bouts, increased marble burying, and increased nose poking in the holeboard assay. MoSeq analysis of behavior further showed signs of altered grooming behaviors and sequencing of behavioral syllables. Cntnap2-/- mice also displayed cognitive inflexibility in a four-choice odor-based reversal learning assay. While they performed similarly to WT controls during acquisition and recall phases, they required significantly more trials to learn a new odor-reward association during reversal, consistent with potential deficits in corticostriatal function.

Strengths:

This study provides significant contributions to the field. The finding of altered SPN excitability, the detailed characterization of striatal inhibition, and the comprehensive behavioral analysis are novel and valuable to understanding the pathophysiology of Cntnap2-/- mice.

Weaknesses:

(1) The approach based on Thy-ChR2 mice has the advantage of overcoming issues caused by injection efficiency and targeting variability. However, the spread of oEPSC amplitudes across mice shown in panels of Figure 1 G/I is very high with almost one order of magnitude difference between some mice. Given this is one of the most important points of the study it will be important to further analyze and discuss what this variability might be due to. Typically, in acute slice recordings, the within-animal variability is larger than the variability across animals. From the sample sizes reported it seems the authors sampled a large number of animals, but with a relatively low number of neurons per animal (per condition). Could this be one of the reasons for this variability?

(2) This is particularly important because the analysis of corticostriatal evoked APs in panels C and E is performed on pooled data without considering the variability in evoked current amplitudes across animals shown in G and I. Were the neurons in panels C/E recorded from the same mice as shown in G/I? If so, it would be informative to regress AP firing data (say at 20% LED) to the average oEPSC amplitude recorded on those mice at the same light intensity. However, if the low number of neurons recorded per mouse is due to technical limitations, then increasing the sample size of these experiments would strengthen the study.

(3) On a similar note, there is no discussion of why iSPNs also show increased corticostriatal evoked firing in Figure 1E, despite the difference in intrinsic excitability shown in Figure 3. This suggests other potential mechanisms that might underlie altered corticostriatal responses. Given the role of Caspr2 in clustering K channels in axons, altered presynaptic function or excitability could also contribute to this phenotype, but potential changes in PPR have not been explored in this study.

(4) Male and female SPNs have different intrinsic properties but the number and/or balance of M/F mice used for each experiment is not reported.

(5) There is no mention of how membrane resistance was calculated, and no I/V plots are shown.

(6) It would be interesting to see which behavior transitions most contribute to the decrease in entropy. Are these caused by repeated or perseverative grooming bouts? Or is this inflexibility also observed across other behaviors? The transition map in Figure S5 shows the overall number of syllables and transitions but not their sequence during behavior. Can this be analyzed by calculating the ratio of individual 𝑢𝑖 × 𝑝𝑖,𝑗 × log2 𝑝𝑖,𝑗 factors across genotypes?

Reviewer #3 (Public Review):

Summary:

The authors analyzed Cntnap2 KO mice to determine whether loss of the ASD risk gene CNTNAP2 alters the dorsal striatum's function.

Strengths:

The results demonstrate that loss of Cntnap2 results in increased excitability of striatal projection neurons (SPNs) and altered striatal-dependent behaviors, such as repetitive, inflexible behaviors. Unlike other brain areas and cell types, synaptic inputs onto SPNs were normal in Cntnap2 KO mice. The experiments are well-designed, and the results support the authors' conclusions.

Weaknesses:

The mechanism underlying SPN hyperexcitability was not explored, and it is unclear whether this cellular phenotype alone can account for the behavioral alterations in Cntnap2 KO mice. No clear explanation emerges for the variable phenotype in different brain areas and cell types.

Author response:

The following is the authors’ response to the original reviews

We thank the reviewers for their careful review of our manuscript and the constructive comments. We have addressed the majority of comments with either new experiments, analyses, and/or text revisions. A summary of the major changes is listed below, followed by our point-by-point responses to the reviewer comments.

Major changes:

(1) We sought to gain insight into the potential mechanistic cause of the increased intrinsic excitability of Cntnap2-/- dSPNs. Given that Kv1.1 and 1.2 potassium channels are known to interact with Caspr2 (the protein encoded by Cntnap2), we hypothesized that altered number, location, and/or function of these channels may underlie the excitability change in these cells. To investigate this, we performed new analyses of the initial dataset to assess action potential (AP) properties known to be impacted by potassium channel function. Indeed, we found that AP frequency was increased, and rheobase current, AP latency and AP threshold were decreased in Cntnap2-/- dSPNs, suggestive of altered Kv1.2 function. These data are in the new Supplemental Fig. 4. We also performed new electrophysiology experiments in which we pharmacologically blocked Kv1.1 and 1.2 to assess whether the effects of blocking these channels would be occluded in Cntnap2-/- dSPNs. We found that 1) WT dSPNs responded to blockade of Kv1.1/1.2 channels by increasing their excitability but Cntnap2-/- dSPNs did not and 2) Kv1.1/1.2 channels were more important contributors to the excitability of dSPNs compared to iSPNs. These new data are presented in the revised Fig. 4 and Supplemental. Figs. 5 and 6.

(2) We performed additional experiments to assess excitatory synaptic properties, specifically AMPA/NMDA receptor ratio. This has been added to Fig. 1.

(3) We performed more rigorous statistical analyses of the initial physiology datasets to align with the statistics performed for the revision experiments. This applies to Fig. 1, Fig. 2, Fig. 3, Fig. 5, and Supp. Fig. 2.

(4) In the discussion section, we now highlight potential limitations of the study and further discuss the variable impact that Cntnap2 loss has on different cell types and brain regions.

Reviewer #1 (Public Review):

Summary:

Cording et al. investigated how deletion of CNTNAP2, a gene associated with autism spectrum disorder, alters corticostriatal engagement and behavior. Specifically, the authors present slice electrophysiology data showing that striatal projection neurons (SPNs) are more readily driven to fire action potentials in response to stimulation of corticostriatal afferents, and this is due to increases in SPN intrinsic excitability rather than changes in excitatory or inhibitory synaptic inputs. The authors show that CNTNAP2 mice display repetitive behaviors, enhanced motor learning, and cognitive inflexibility. Overall the authors' conclusions are supported by their data, but a few claims could use some more evidence to be convincing.

Strengths:

The use of multiple behavioral techniques, both traditional and cutting-edge machine learning-based analyses, provides a powerful means of assessing repetitive behaviors and behavioral transitions/rigidity.

Characterization of both excitatory and inhibitory synaptic responses in slice electrophysiology experiments offers a broad survey of the synaptic alterations that may lead to increased corticostriatal engagement of SPNs.

Weaknesses:

(1) The authors conclude that increased cortical engagement of SPNs is due to changes in SPN intrinsic excitability rather than synaptic strength (either excitatory or inhibitory). One weakness is that only AMPA receptor-mediated responses were measured. Though the holding potential used for experiments in Figure 1FI wasn't clear, recordings were presumably performed at a hyperpolarized potential that limits NMDA receptormediated responses. Because the input-output experiments used to conclude that corticostriatal engagement of SPNs is elevated (Figure 1B-E) were conducted in the current clamp, it is possible that enhanced NMDA receptor engagement contributed to increased SPN responses to cortical stimulation. Confirming that NMDA receptor-mediated EPSC components are not altered would strengthen the main conclusion.

The reviewer is correct, the initial optically-evoked EPSC assessments were performed at a hyperpolarized potential (-70mV), thus measuring primarily AMPAR-mediated currents. We agree that assessing potential changes in the NMDAR-mediated EPSC component is important and we have completed new experiments to assess this. We find no differences in NMDAR-mediated EPSCs assessed at +40mV or the AMPA:NMDA ratio.

These results have been added to Fig. 1. An expanded analysis of these results is shown in Author response image 1. We note that the previous AMPAR-mediated EPSC results have been replicated in this additional experiment, again showing no change in Cntnap2-/- SPNs.

AMPA and NMDA receptor-mediated EPSCs are unchanged in Cntnap2-/- SPNs.

(A) Quantification (mean ± SEM) of AMPA:NMDA ratio per cell for Cntnap2+/+ and Cntnap2-/- dSPNs, p=0.9537, MannWhitney test. (B) dSPN AMPA current per cell, p=0.6172, Mann-Whitney test. (C) dSPN NMDA current per cell, p=0.6009, Mann-Whitney test. (D) dSPN AMPA:NMDA ratio averaged by animal, p=0.8413, Mann-Whitney test. (E) dSPN AMPA current averaged by animal, p>0.9999, Mann-Whitney test. (F) dSPN NMDA current averaged by animal, p=0.6905, Mann-Whitney test. (G) Quantification (mean ± SEM) of AMPA:NMDA ratio per cell for Cntnap2+/+ and Cntnap2-/- iSPNs, p=0.4104, Mann-Whitney test. (H) iSPN AMPA current per cell, p=0.9010, Mann-Whitney test. (I) iSPN NMDA current per cell, p=0.9512, two-tailed unpaired t test. (J) iSPN AMPA:NMDA averaged by animal, p=0.3095, Mann-Whitney test. (K) iSPN AMPA current averaged by animal, p=>0.9999, Mann-Whitney test. (L) iSPN NMDA current averaged by animal, p=0.8413, MannWhitney test. All values were recorded using 20% blue light intensity. For dSPNs: Cntnap2+/+ n=22 cells from 5 mice, Cntnap2-/- n=22 cells from 5 mice. For iSPNs: Cntnap2+/+ n=21 cells from 5 mice, Cntnap2-/-n=21 cells from 5 mice.

(2) Data clearly show that SPN intrinsic excitability is increased in knockout mice. Given that CNTNAP2 has been linked to potassium channel regulation, it would be helpful to show and quantify additional related electrophysiology data such as negative IV curve responses and action potential hyperpolarization.

We appreciate this suggestion. As indicated by the reviewer, Caspr2, has previously been shown to control the clustering of Kv1-family potassium channels in axons isolated from optic nerve and corpus callosum (PMIDs: 10624965, 12963709, 29300891). In particular, Caspr2 is known to associate directly with Kv1.2 (PMID: 29300891). To assess a potential contribution of Kv1.2 to the excitability phenotype, we performed additional analyses of our original dataset to quantify AP properties known to be impacted by changes in Kv1.2 function (i.e. latency to fire and AP threshold, new Supp. Fig. 4). We identified several changes in Cntnap2-/- dSPNs resembling those that occur in wild-type cells when Kv1.2 is blocked (i.e. reduced threshold and reduced latency to fire, Supp. Fig. 4).

We then performed a pharmacological experiment, blocking Kv1.2 using α-dendrotoxin (α-DTX) while recording intrinsic excitability to assess whether the effects of this drug on dSPN excitability were occluded in Cntnap2-/- cells. Indeed, we found that while blocking Kv1.2 in wild-type dSPNs significantly reduced threshold and increased intrinsic excitability, these effects were not seen in Cntnap2-/- dSPNs (new Fig. 4). We believe that this suggests an altered contribution of Kv1.2 to the intrinsic excitability of mutant dSPNs, owing to a change in the clustering, number, or function of these channels. Therefore, loss-of-function of Kv1.2 is a likely explanation for the enhanced intrinsic excitability of Cntnap2-/- dSPNs. Interestingly, we found that α-DTX had only subtle effects on iSPNs (Cntnap2 WT or mutant), suggesting a lesser contribution of this channel in controlling the excitability of indirect pathway cells. This finding can account for the relatively stronger effect of Cntnap2 loss on dSPN physiology. The results of these new experiments and analyses are presented in the new Fig. 4, Supp. Fig. 5 and Supp. Fig. 6.

(3) As it stands, the reported changes in dorsolateral striatum SPN excitability are only correlative with reported changes in repetitive behaviors, motor learning, and cognitive flexibility.

We agree that we have not identified a causative relationship between the change in dorsolateral dSPN excitability and the behaviors that we measured in Cntnap2-/- mice. That said, in a previous study, we showed that selective deletion of the autism spectrum disorder (ASD) risk gene Tsc1 from dorsal striatal dSPNs resulted in increased corticostriatal drive and this was sufficient to increase rotarod motor learning (PMID: 34380034). Therefore, while we have not demonstrated causality in this study, we hypothesize that changes in dSPN excitability are likely to contribute to the behavioral phenotypes observed in Cntnap2-/- mice.

Reviewer #2 (Public Review):

Summary:

This is an important study characterizing striatal dysfunction and behavioral deficits in Cntnap2-/- mice. There is growing evidence suggesting that striatal dysfunction underlies core symptoms of ASD but the specific cellular and circuit level abnormalities disrupted by different risk genes remain unclear. This study addresses how the deletion of Cntnap2 affects the intrinsic properties and synaptic connectivity of striatal spiny projection neurons (SPN) of the direct (dSPN) and indirect (iSPN) pathways. Using Thy1-ChR2 mice and optogenetics the authors found increased firing of both types of SPNs in response to cortical afferent stimulation. However, there was no significant difference in the amplitude of optically-evoked excitatory postsynaptic currents (EPSCs) or spine density between Cntnap2-/- and WT SPNs, suggesting that the increased corticostriatal coupling might be due to changes in intrinsic excitability. Indeed, the authors found Cntnap2-/- SPNs, particularly dSPNs, exhibited higher intrinsic excitability, reduced rheobase current, and increased membrane resistance compared to WT SPNs. The enhanced spiking probability in Cntnap2-/- SPNs is not due to reduced inhibition. Despite previous reports of decreased parvalbumin-expressing (PV) interneurons in various brain regions of Cntnap2-/- mice, the number and function (IPSC amplitude and intrinsic excitability) of these interneurons in the striatum were comparable to WT controls.

This study also includes a comprehensive behavioral analysis of striatal-related behaviors. Cntnap2-/- mice demonstrated increased repetitive behaviors (RRBs), including more grooming bouts, increased marble burying, and increased nose poking in the holeboard assay. MoSeq analysis of behavior further showed signs of altered grooming behaviors and sequencing of behavioral syllables. Cntnap2-/- mice also displayed cognitive inflexibility in a four-choice odor-based reversal learning assay. While they performed similarly to WT controls during acquisition and recall phases, they required significantly more trials to learn a new odor-reward association during reversal, consistent with potential deficits in corticostriatal function.

Strengths:

This study provides significant contributions to the field. The finding of altered SPN excitability, the detailed characterization of striatal inhibition, and the comprehensive behavioral analysis are novel and valuable to understanding the pathophysiology of Cntnap2-/- mice.

Weaknesses:

(1) The approach based on Thy-ChR2 mice has the advantage of overcoming issues caused by injection efficiency and targeting variability. However, the spread of oEPSC amplitudes across mice shown in panels of Figure 1 G/I is very high with almost one order of magnitude difference between some mice. Given this is one of the most important points of the study it will be important to further analyze and discuss what this variability might be due to. Typically, in acute slice recordings, the within-animal variability is larger than the variability across animals. From the sample sizes reported it seems the authors sampled a large number of animals, but with a relatively low number of neurons per animal (per condition). Could this be one of the reasons for this variability?

We agree with the reviewer that the variability in these experiments is quite large. We have replicated these experiments in the process of performing AMPA:NMDA ratio recordings (see above response to Reviewer 1’s comment). We again find no differences in AMPAR-mediated EPSC amplitude between WT and mutant SPNs (Author response image 2). Notably, these experiments also demonstrate a large amount of variability. In the original dataset, a small number of cells were collected from each animal (~1-3 cells/mouse). However, the variability remains in the new dataset, in which more cells were collected from each animal (~4-6 cells/mouse). We find both withinanimal and between-animal variability, as can be seen in Author response image 2 (recordings made from the same animal are color-coordinated). Potential sources of variability in this experiment include: 1) variable expression of ChR2 per mouse, 2) variable innervation of ChR2-expressing terminals onto any given recorded cell, and/or 3) differences in prior plasticity state between cells (i.e. some neurons may have recently undergone corticostriatal LTP or LTD).

Author response image 2.

Optically-evoked AMPAR EPSCs exhibit within- and between-animal variability.

(A) Quantification of EPSC amplitude evoked in dSPNs at different light intensities from the original dataset, plotted by cell (line represents the mean, dots/squares represent average EPSC amplitude for each recorded cell). Cntnap2+/+ n=17 cells from 8 mice, Cntnap2-/- n=13 cells from 5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 56) = 0.3879, geno F (1, 28) = 0.8098, stim F (1.047, 29.32) = 76.56. (B) Quantification of EPSC amplitude evoked in dSPNs, averaged by mouse (line represents the mean, dots/squares represent average EPSC amplitude for each mouse). Cntnap2+/+ n=8 mice, Cntnap2-/- n=5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 22) = 0.2154, geno F (1, 11) = 0.2585, stim F (1.053, 11.58) = 49.68. (C) Quantification of EPSC amplitude in dSPNs from the revision dataset, plotted by cell (line represents the mean, dots/squares represent average EPSC amplitude for each recorded cell). Cntnap2+/+ n=22 cells from 5 mice, Cntnap2-/- n=22 cells from 5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 84) = 0.01885, geno F (1, 42) = 0.002732, stim F (1.863, 78.26) = 20.93. (D) Quantification of EPSC amplitude in dSPNs from the revision dataset, averaged by mouse (line represents the mean, dots/squares represent average EPSC amplitude for each mouse). Cntnap2+/+ n=5 mice, Cntnap2-/- n=5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 16) = 0.06288, geno F (1, 8) = 0.006548, stim F (1.585, 12.68) = 16.97. (E) Quantification of EPSC amplitude evoked in iSPNs from the original dataset, plotted by cell (line represents the mean, dots/squares represent average EPSC amplitude for each recorded cell). Cntnap2+/+ n=13 cells from 6 mice, Cntnap2-/- n=11 cells from 5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 44) = 0.9414, geno F (1, 22) = 1.333, stim F (1.099, 24.18) = 52.26. (F) Quantification of EPSC amplitude evoked in iSPNs from original dataset, averaged by mouse (line represents the mean, dots/squares represent average EPSC amplitude for each mouse). Cntnap2+/+ n=6 mice, Cntnap2-/- n=5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 18) = 0.4428, geno F (1, 9) = 0.5635, stim F (1.095, 9.851) = 23.82. (G) Quantification of EPSC amplitude evoked in iSPNs from the revision dataset, plotted by cell (line represents the mean, dots/squares represent average EPSC amplitude for each recorded cell). Cntnap2+/+ n=21 cells from 5 mice, Cntnap2-/- n=21 cells from 5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 80) = 0.04134, geno F (1, 40) = 0.007025, stim F (1.208, 48.31) = 102.9. (H) Quantification of EPSC amplitude evoked in iSPNs from the revision dataset, averaged by mouse (line represents the mean, dots/squares represent average EPSC amplitude for each mouse). Cntnap2+/+ n=5 mice, Cntnap2-/- n=5 mice. Repeated measures two-way ANOVA p values are shown; g x s F (2, 16) = 0.001865, geno F (1, 8) = 0.1004, stim F (1.179, 9.433) = 61.31.

(2) This is particularly important because the analysis of corticostriatal evoked APs in panels C and E is performed on pooled data without considering the variability in evoked current amplitudes across animals shown in G and I. Were the neurons in panels C/E recorded from the same mice as shown in G/I? If so, it would be informative to regress AP firing data (say at 20% LED) to the average oEPSC amplitude recorded on those mice at the same light intensity. However, if the low number of neurons recorded per mouse is due to technical limitations, then increasing the sample size of these experiments would strengthen the study.

We appreciate this point; however, the evoked AP experiment and the evoked EPSC experiment were performed on different mice, so it is not possible to correlate the data across experiments. While the evoked AP experiments were performed using potassium-based internal, we used a cesium-based internal to measure AMPAR-mediated EPSCs to more accurately detect synaptic currents. We note that the evoked AP experiments share a similar amount of variability as the evoked EPSC experiments, again possibly owing to variable expression of channelrhodopsin per mouse, variable innervation of ChR2-positive terminals onto individual cells, and/or differences in prior plasticity status between cells.

(3) On a similar note, there is no discussion of why iSPNs also show increased corticostriatal evoked firing in Figure 1E, despite the difference in intrinsic excitability shown in Figure 3. This suggests other potential mechanisms that might underlie altered corticostriatal responses. Given the role of Caspr2 in clustering K channels in axons, altered presynaptic function or excitability could also contribute to this phenotype, but potential changes in PPR have not been explored in this study.

We have now performed more rigorous statistics on the data in Fig. 1 (repeated measures two-way ANOVA) such that the difference in corticostriatal evoked firing in Cntnap2-/- iSPNs no longer reaches statistical significance. This is consistent with the modest but statistically non-significant effect of Cntnap2 loss on iSPN intrinsic excitability. We agree with the reviewer that presynaptic alterations could potentially contribute to the changes in cortically-driven action potentials, especially as this experiment was performed without any synaptic blockers present, and Cntnap2 is deleted from all cells. That said, if changes in presynaptic release probability accounted for the increased corticostriatal drive, we would expect to see differences in cortically-evoked EPSCs onto SPNs.

While we can’t rule out the possibility of pre-synaptic changes, a straightforward explanation for our findings is that loss or alteration of Kv1.2 channel function is responsible for the increased excitability of Cntnap2-/- dSPNs, resulting in enhanced spiking in response to cortical input. Given the fact that Kv1.2 channels appear less important for regulating iSPN excitability (see new Fig. 4 and Supp. Fig. 6), this can explain the greater impact of Cntnap2 loss on dSPN physiology.

(4) Male and female SPNs have different intrinsic properties but the number and/or balance of M/F mice used for each experiment is not reported.

We agree that this is an important consideration. Author response table 1 provides the sex breakdown for the intrinsic excitability experiments. While we did not explicitly power the experiments to test for sex differences, Author response image 3 shows the data separated by sex and genotype for the intrinsic excitability experiments. Within genotype, we find no significant differences between males and females, except for Cntnap2-/- iSPNs which showed a significant interaction between sex and current step (Author response image 3F). Interestingly, while present in both sexes, the excitability shift of Cntnap2-/- dSPNs may be slightly more pronounced in females compared to males (Author response image 3C and D). However, this result would require further validation with a greater sample size.

Enhanced excitability of Cntnap2-/- dSPNs is present in both males and females.

(A) Quantification (mean ± SEM) of the number of APs evoked in dSPNs in Cntnap2+/+ males and females at different current step amplitudes. Cntnap2+/+ males n=12 cells from 4 mice, Cntnap2+/+ females n=8 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; s x c F (28, 560) = 0.8992, sex F (1, 20) = 0.3754, current F (1.279, 25.57) = 56.85. (B) Quantification (mean ± SEM) of the number of APs evoked in dSPNs in Cntnap2-/- males and females at different current step amplitudes. Cntnap2-/- males n=12 cells from 4 mice, Cntnap2-/- females n=11 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; s x c F (28, 588) = 0.6752, sex F (1, 21) = 0.04534, current F (2.198, 46.15) = 78.89. (C) Quantification (mean ± SEM) of the number of APs evoked in dSPNs in Cntnap2+/+ males and Cntnap2-/- males at different current step amplitudes. Cntnap2+/+ males n=12 cells from 4 mice, Cntnap2-/- males n=12 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; g x c F (28, 672) = 2.233, geno F (1, 24) = 3.746, current F (1.708, 40.98) = 79.82. (D) Quantification (mean ± SEM) of the number of APs evoked in dSPNs in Cntnap2+/+ females and Cntnap2-/- females at different current step amplitudes. Cntnap2+/+ females n=8 cells from 4 mice, Cntnap2-/- females n=11 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; g x c F (28, 476) = 1.547, geno F (1, 17) = 5.912, current F (1.892, 32.17) = 58.76. (E) Quantification (mean ± SEM) of the number of APs evoked in iSPNs in Cntnap2+/+ males and females at different current step amplitudes. Cntnap2+/+ males n=10 cells from 4 mice, Cntnap2+/+ females n=12 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; s x c F (28, 560) = 1.236, sex F (1, 20) = 1.074, current F (2.217, 44.34) = 179.6. (F) Quantification (mean ± SEM) of the number of APs evoked in iSPNs in Cntnap2-/- males and females at different current step amplitudes. Cntnap2-/- males n=12 cells from 4 mice, Cntnap2-/- females n=9 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; s x c F (28, 532) = 2.513, sex F (1, 19) = 2.639, current F (1.858, 35.31) = 152.5. (G) Quantification (mean ± SEM) of the number of APs evoked in iSPNs in Cntnap2+/+ males and Cntnap2-/- males at different current step amplitudes. Cntnap2+/+ males n=10 cells from 4 mice, Cntnap2-/- males n=12 cells from 4 mice. Repeated measures twoway ANOVA p values are shown; g x c F (28, 560) = 0.4723, geno F (1, 20) = 0.5675, current F (2.423, 48.47) = 301.7. (H) Quantification (mean ± SEM) of the number of APs evoked in iSPNs in Cntnap2+/+ females and Cntnap2-/- females at different current step amplitudes. Cntnap2+/+ females n=12 cells from 4 mice, Cntnap2-/- females n=9 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; g x c F (28, 532) = 1.655, geno F (1, 19) = 0.2322, current F (2.081, 39.55) = 99.45.

(5) There is no mention of how membrane resistance was calculated, and no I/V plots are shown.

Passive properties were calculated from the average of five -5 mV, 100 ms long test pulse steps applied at the beginning of every experiment. Membrane resistance was calculated from the double exponential curve fit. This has now been added to the methods section.

(6) It would be interesting to see which behavior transitions most contribute to the decrease in entropy. Are these caused by repeated or perseverative grooming bouts? Or is this inflexibility also observed across other behaviors? The transition map in Figure S5 shows the overall number of syllables and transitions but not their sequence during behavior. Can this be analyzed by calculating the ratio of individual 𝑢𝑖 × 𝑝𝑖,𝑗 × log2 𝑝𝑖,𝑗 factors across genotypes?

We thank the reviewer for raising an insightful question. Here we use a finite state Markov chain model to describe the syllable transitions in animal behavior. To quantify the randomness in the system, we calculated the entropy of the Markov chain (see methods section). The reviewer suggested calculating the partial entropy of the transition matrix, which would allow us to estimate the contribution of a subset of states to the entropy of the whole system, given by the equation:

The partial equation can indeed quantify the stochasticity, or “flexibility” in our context, of the sub-system containing only a subset of the behavior syllables. However, there are two main limitations to this approach:

(1) The partial entropy fails to account for the transitions connecting the subset with the rest of the states in the system

(2) The stationary distribution may not reflect the actual probabilities in the isolated sub-system S.

Consequently, the partial entropy cannot be directly interpreted as the fraction of contributions from specific syllable pairs or sub-system to the entropy of the whole system. To be more specific, while a significant difference between the same sub-system in WT and KO groups could indicate that the sub-system contributes significantly to the difference of overall entropy, a non-significant result does not mean that the sub-system does not contribute to overall entropy difference, as interactions between the sub-system and other notconsidered states are not accounted for.

Grooming syllables contribute to some but not all differences in syllable transitions in Cntnap2-/- mice.

We calculated the entropy of each syllable pair using 𝑢𝑖 × 𝑝𝑖,𝑗 × log2 𝑝𝑖,𝑗 for every syllable pair and every animal. We then statistically tested the difference between genotypes for each syllable pair using Mann-Whitney tests. This plot displays those adjusted p-values for each syllable pair between WT and KO groups. The significant p-values suggest that the transitions to syllables 24 and 25 are different between genotypes (note that these correspond to grooming syllables, see Fig. 5N). However, since the overall entropy is a summation of every pair, it is difficult to conclude that syllables 24 and 25 are the sole contributors to the different entropy we observed.

Reviewer #3 (Public Review):

Summary:

The authors analyzed Cntnap2 KO mice to determine whether loss of the ASD risk gene CNTNAP2 alters the dorsal striatum's function.

Strengths:

The results demonstrate that loss of Cntnap2 results in increased excitability of striatal projection neurons (SPNs) and altered striatal-dependent behaviors, such as repetitive, inflexible behaviors. Unlike other brain areas and cell types, synaptic inputs onto SPNs were normal in Cntnap2 KO mice. The experiments are welldesigned, and the results support the authors' conclusions.

Weaknesses:

The mechanism underlying SPN hyperexcitability was not explored, and it is unclear whether this cellular phenotype alone can account for the behavioral alterations in Cntnap2 KO mice. No clear explanation emerges for the variable phenotype in different brain areas and cell types.

We agree that identifying the mechanism by which Cntnap2 loss affects intrinsic excitability is interesting and important. We have added experiments to address this and conclude that the improper clustering, number, or function of Kv1.2 channels in Cntnap2-/- dSPNs is likely responsible for their increased excitability. These channels are known to be clustered/organized in part by Caspr2 (PMIDs: 10624965, 12963709, 29300891), and Kv1.2 channels are known to play an important role in regulating excitability in SPNs (PMIDs: 13679409, 32075716). In the case of dSPNs, blocking these channels with α-DTX significantly increased the excitability of WT cells (as has been previously reported); however, this effect was occluded in mutant cells, perhaps owing to a decreased contribution of Kv1.2 channels to excitability in Cntnap2-/- dSPNs. In addition, we found that blockade of these channels with α-DTX only modestly affected the excitability of iSPNs. Therefore, this can explain why loss of Cntnap2 more strongly affects the excitability of dSPNs. Please see new Fig. 4, Supp. Fig. 5 and Supp. Fig. 6 for these new data.

We agree with the reviewer that we have not identified a causative relationship between the change in dSPN excitability and the behavioral alterations in Cntnap2-/- mice. This is a limitation of the study.

It is interesting to speculate on the root of the varying impacts to excitability that occur across different brain regions and cell types in Cntnap2-/- mice. Increased excitability, as we see in dSPNs, has been identified in cerebellar Purkinje cells and L2/3 pyramidal neurons in somatosensory cortex in the context of Cntnap2 loss (PMIDs: 34593517, 30679017, 36793543). However, other cell types in Cntnap2-/- mice have exhibited no change in excitability (mPFC, L2/3 pyramidal neurons, PMID: 31141683) or hypoexcitability (subset of L5/6 pyramidal neurons, PMID: 29112191). While all of these cell types express Kv1.2 channels, they fundamentally vary in their intrinsic properties, owing to the role that other ion channels play in membrane excitability. As a result, loss of Cntnap2 is expected to have a variable effect on excitability depending on the cell type and the complement of other ion channels that are present. In addition, an initial change in excitability may drive secondary, potentially compensatory, changes in other channels that lead to a different excitability state. These changes are also expected to be cell type-specific. We do note that both of the cell types that show increased excitability in the context of Cntnap2 loss have been shown to exhibit an α-DTX-sensitive Kv1 channel current, such that application of α-DTX results in increased firing of these cells (cerebellar Purkinje cells; PMIDs: 17087603, 16210348 and L2/3 pyramidal neurons in somatosensory cortex; PMID: 17215507). These findings are consistent with our results in Cntnap2-/- dSPNs.

Reviewer #1 (Recommendations For The Authors):

More thorough analysis of some of the manually quantified behaviors would be helpful. For example, only the grooming bout number was presented- what about the duration of bouts and total time grooming? Similarly, for the open field the number of center entries was reported but what about the total time in the center?

We have quantified the time spent grooming and total time spent in the center during the open field test from our original data (Author response image 5). These data were not originally included in the manuscript because they were recorded for only a subset of the total animals. For each of these measures we find trend level changes, which are consistent with the primary measures reported in the main manuscript.

Time in center and time spent grooming trend towards an increase in Cntnap2-/- mice.

(A) Quantification (mean ± SEM) of total time spent in the center of the open field during a 60 minute test, p=0.0656, Mann-Whitney test. (B) Time spent grooming during the first 20 minutes of the open field test, p=0.0611, Mann-Whitney test. For both measurements, Cntnap2+/+ n=18 mice, Cntnap2-/- n=19 mice.

Reviewer #3 (Recommendations For The Authors):

What accounts for the hyperexcitability observed in Cntnap2-deficient SPNs? The authors noted that excitability is reportedly increased, reduced, or unchanged in different brain areas. What accounts for this disparity? Is it about the subcellular localization of Kv1 channels? The authors may want to test this possibility experimentally. At least, they may want to test whether Kv1 channels are mislocalized in SPNs.

We agree that this is an important point, and we have performed additional experiments to address this. We find that the Kv1.2 blocker a-DTX significantly increases the excitability of WT dSPNs but not Cntnap2-/- dSPNs. This suggests that the mechanism underlying dSPN hyperexcitability in Cntnap2 mutants is the improper clustering, number, or function of Kv1.2 channels. These channels are known to be clustered and organized in part by Caspr2 (PMIDs: 10624965, 12963709, 29300891) and have been shown to play an important role in regulating the excitability of SPNs (PMIDs: 13679409, 32075716). Interestingly, we find that a-DTX has less of an effect on the excitability of iSPNs, which may account for the greater impact of Cntnap2 loss on dSPNs. Please see new Fig. 4, Supp. Fig. 5 and Supp. Fig. 6 for these added data and analyses.

Please see above response to Reviewer #3 for our speculation on the variable impact of Cntnap2 loss on different cell types and brain regions.

We agree with the reviewer that assessing potential differences in subcellular localization of Kv1 channels in our model would bolster the conclusion that these channels are mislocalized in the Cntnap2-/- striatum. We piloted these experiments using immunohistochemistry to stain for Kv1.1 and 1.2 but found that without very high-resolution imaging, it would be challenging to accurately quantify Kv1 puncta in a cell type-specific manner. We instead chose to investigate the functional contribution of Kv1 channels to the dSPN hyperexcitability phenotype through the a-DTX experiments outlined above. α-DTX strongly inhibits Kv1.2 channels, but also Kv1.1 channels to some extent (PMIDs: 12042352, 13679409). We find that the effects of a-DTX on SPN excitability are occluded in Cntnap2-/- dSPNs; therefore, we conclude that Kv1.2 (and possibly Kv1.1) channels have reduced function in these cells. Further work will be needed to determine if this is a result of channel mislocalization or another type of alteration.

The authors did not detect synaptic changes in Cntnap-deficient SPNs. This important observation should be briefly discussed in the context of previous work in other brain regions and cell types. For example, some studies reported structural and functional changes at excitatory synapses. The variable impact on synapses suggests distinct compensatory mechanisms in different brain areas.

Given the prior literature showing effects of Cntnap2 loss on synapses in other brain regions, we were surprised that striatal synapses were not impacted in our model. We agree with the reviewer that the variable changes in synaptic properties across brain regions in Cntnap2 mutant mice is likely a result of distinct compensatory changes in these regions. Differences may also arise depending on whether the synaptic changes originate from the post-synaptic cell or from pre-synaptic changes. An interesting direction for future studies would be to explore the developmental trajectory of excitability and synaptic changes to determine which may be initial perturbations versus those that are secondary and potentially compensatory.

Line 138: "synaptic excitability". How is this term defined? Consider "synaptic changes" instead.

“Synaptic excitability” was used to mean a change in the number and/or function of glutamate receptors. We have now changed this term to “excitatory synaptic changes.”

Consider a short paragraph to highlight some limitations of this study. For example, it is unclear whether SPN hyperexcitability results from a compensatory change in Cntnap2 KO mice and whether the behavioral phenotype is solely due to this cellular phenotype. The study focuses on cortical projections onto SPNs, but these cells receive inputs from other brain areas that were not explored. Lastly, no clear explanation emerges for the variable phenotype in different brain areas and cell types.

We thank the reviewer for this suggestion and have added several paragraphs to the discussion highlighting some limitations of this study.

We hypothesize that the dSPN hyperexcitability in Cntnap2-/- mice is a primary change, due to the direct relationship between Caspr2 and Kv1.2 channels. The results of our -DTX experiments suggest that the function and/or contribution of these channels to excitability is altered in Cntnap2-/- dSPNs. However, it is possible that there are additional changes in dSPNs that occur as a result of Cntnap2 loss and contribute to the hyperexcitability of these cells. Rather surprisingly, we don’t find evidence for altered excitatory (specifically from cortical inputs) or inhibitory synaptic function, suggesting lack of engagement of homeostatic mechanisms at the synaptic level.

We have not yet determined whether there is a causative relationship between the change in dSPN excitability and the behavioral alterations in Cntnap2-/- mice. This is a limitation of the current study. In our discussion section, we highlight that the dSPN changes we observe in dorsolateral striatum (DLS) are known to be sufficient to enhance rotarod learning in other mouse models and thus supports a connection between this cellular change and behavior. For the other behaviors we measured, we acknowledge that both DLS and other striatal or extra-striatal brain regions have been implicated in these behaviors, and therefore less of a direct connection can be made.

In terms of the inputs, we focused on cortical inputs given their known role in mediating motor and habit learning (PMID: 15242609, 16237445, 19198605). Notably, corticostriatal synapses have been shown to be altered across a variety of mouse models with mutations in ASD risk genes and therefore may be a point of convergence for disparate genetic insults (PMID: 31758607). We agree that the striatum receives inputs from a variety of brain regions, notably the thalamus, which we did not explore in this study. This would be an interesting area for future studies.

Finally, it is difficult to speculate on the root of the varying impacts to excitability that occur across different brain regions and cell types in Cntnap2-/- mice. Please see above response to Reviewer #3 for some speculation on this point in regard to the potential involvement of Kv1.2 in the excitability changes in various Cntnap2-/- cell types. To expand upon this, it is known that ASD-associated mutations can have varying impacts on cell function even across similar cell types within a given brain region – we have seen this between dSPNs and iSPNs (this study, PMIDs: 34380034, 39358043), as have other groups studying ASD risk gene mutations in striatum (PMID: 24995986). This differential impact of the same mutation on intrinsic and/or synaptic physiology across cell types has been identified in other brain regions as well (PMID: 22884327, 26601124). Differences in transcriptional programs, protein expression, neuronal morphology, synaptic inputs and plasticity state make up a non-exhaustive set of variables that will impact the physiological function of a neuron, both in terms of the direct but also indirect consequences of an ASD risk gene mutation. To better address this important question, future studies would benefit from a systematic approach to assessing physiological changes in a given ASD mouse model, both across development and across brain regions.

Version 2

Abstract

Abstract

Autism spectrum disorder (ASD) is a neurodevelopmental disorder characterized by two major diagnostic criteria – persistent deficits in social communication and interaction, and the presence of restricted, repetitive patterns of behavior (RRBs). Evidence from both human and animal model studies of ASD suggest that alteration of striatal circuits, which mediate motor learning, action selection, and habit formation, may contribute to the manifestation of RRBs. CNTNAP2 is a syndromic ASD risk gene, and loss of function of Cntnap2 in mice is associated with RRBs. How loss of Cntnap2 impacts striatal neuron function is largely unknown. In this study, we utilized Cntnap2−/− mice to test whether altered striatal neuron activity contributes to aberrant motor behaviors relevant to ASD. We find that Cntnap2−/− mice exhibit increased cortical drive of direct pathway striatal projection neurons (dSPNs). This enhanced drive is likely due to increased intrinsic excitability of dSPNs, which make them more responsive to cortical inputs. We find that Cntnap2−/− mice exhibit spontaneous repetitive behaviors, increased motor routine learning, perseveration, and cognitive inflexibility. Increased corticostriatal drive of the direct pathway may therefore contribute to the acquisition of repetitive, inflexible behaviors in Cntnap2 mice.

Introduction

Autism spectrum disorder (ASD) is characterized by alterations in social communication and interaction, as well as the presence of restricted, repetitive, inflexible behaviors (APA, 2022). Given that ASD has high heritability (Sandin et al., 2017), much work has been done in the last 30 years to identify genes that confer risk of developing ASD (De Rubeis et al., 2014; Iossifov et al., 2014; Sanders et al., 2015). Through this, hundreds of high-confidence risk genes have been identified, varying greatly in the proteins for which they code (Satterstrom et al., 2020). These include transcriptional and translational regulators, ion channels, receptors, cell adhesion molecules, and others (De Rubeis et al., 2014; Delorme et al., 2013; Ebert & Greenberg, 2013). Recent work has focused on identifying brain regions and circuits that may be commonly affected by ASD-related mutations. The basal ganglia, in particular the striatum, represents one such commonly altered brain region and prior studies have demonstrated changes in striatal function and striatum-associated behaviors in mice with mutations in ASD risk genes (Benthall et al., 2021; Fuccillo, 2016; Peca et al., 2011; Peixoto et al., 2016; Platt et al., 2017; Rothwell et al., 2014; Wang et al., 2016). However, whether basal ganglia circuits are convergently changed in ASD mouse models is an open question. Here we investigated whether loss of function of the syndromic ASD risk gene Cntnap2 alters striatal physiology and basal ganglia-dependent behaviors in mice.

Cntnap2 codes for a neurexin-like cell adhesion molecule called Contactin-associated protein-like 2 (Caspr2) (Poliak et al., 1999; Poliak et al., 2003). In mice, Caspr2 is expressed in several cortical and subcortical regions, including the striatum, from embryonic day 14 (E14) onward into adulthood (Penagarikano et al., 2011). Caspr2 is primarily localized at the juxtaparanodes of myelinated axons and is involved in the clustering of potassium channels (Poliak et al., 2003; Scott et al., 2019). In mice, in vitro studies suggests that Caspr2 may also play a role in AMPAR trafficking and cell morphology (Anderson et al., 2012; Gdalyahu et al., 2015; Varea et al., 2015), and ex vivo experiments indicate that it can control cell excitability and circuit synchronicity (Martin-de-Saavedra et al., 2022). Caspr2 is important during neurodevelopment and has been implicated in neuronal migration (Penagarikano et al., 2011), the maturation and function of parvalbumin-positive GABAergic interneurons (Penagarikano et al., 2011; Scott et al., 2019; Vogt et al., 2018), and the timing of myelination (Scott et al., 2019). CNTNAP2 mutations in people lead to a neurodevelopmental syndrome that can include language disorders, epilepsy, obsessive compulsive disorder, and ASD (Penagarikano & Geschwind, 2012; Rodenas-Cuadrado et al., 2014). A mouse model of this syndrome, Cntnap2−/−, has been shown to exhibit good face validity for ASD-relevant social and motor behavior alterations (Brunner et al., 2015; Dawes et al., 2018; Penagarikano et al., 2011; Scott et al., 2019). However, the impact of Cntnap2 loss on striatal physiology and corticostriatal-dependent behaviors has not been comprehensively assessed.

The striatum is primarily composed of GABAergic striatal projection neurons (SPNs), which make up two functionally distinct output pathways: the D1-receptor expressing cells of the direct pathway (dSPNs), which project to substantia nigra pars reticulata (SNr), and the D2-receptor expressing cells of the indirect pathway (iSPNs), which project to external globus pallidus (GPe) (Calabresi et al., 2014; Gerfen & Surmeier, 2011; Kravitz et al., 2010; Tai et al., 2012). The two types of SPNs are intermixed throughout the striatum and receive excitatory glutamatergic inputs from cortex and thalamus, as well as dopaminergic input from the midbrain (Ding et al., 2008; Doig et al., 2010; Gerfen & Surmeier, 2011). Coordinated activity between the populations of SPNs in response to these inputs mediates action selection, motor learning and habit formation (Hawes et al., 2015; Santos et al., 2015; Yin et al., 2005, 2006). Although SPNs comprise upwards of 95% of the cells in striatum, there are distinct types of GABAergic interneurons that contribute significantly to the inhibitory circuitry of the striatum. Parvalbumin (PV) interneurons, which make up ∼2% of the cells in striatum, provide the largest feedforward inhibition onto SPNs (Burke et al., 2017). Changes in the number and/or function of PV interneurons have been identified in several ASD mouse models, including Cntnap2−/− mice, implicating PV circuitry as a potential common alteration across ASD mouse models (Filice et al., 2020; Juarez & Martinez Cerdeno, 2022).

To determine how loss of Cntnap2 affects striatal function, we assessed the physiology of SPNs and PV-interneurons in the dorsolateral striatum and utilized a range of assays to assess striatum-associated behaviors in Cntnap2−/− mice. We find that SPNs of the direct pathway exhibit increased corticostriatal drive, despite unchanged excitatory cortical input. Although decreased inhibitory function has been identified in other brain regions in Cntnap2−/− mice, we find no deficit in broad or PV-specific inhibitory input onto either SPN subtype in the case of Cntnap2 loss. Instead, we identify a significant increase in the intrinsic excitability of dSPNs in Cntnap2−/− mice, driven by a reduction in Kv1.2 channel function. Behaviorally, we find that Cntnap2−/− mice exhibit RRB-like behaviors including increased grooming, marble burying and nose poking in the holeboard assay. Cntnap2−/− mice also exhibit increased motor routine learning in the accelerating rotarod, and cognitive inflexibility in an odor-based reversal learning task. Taken together, these findings suggest that enhanced direct pathway drive may play a role in the spontaneous and learned repetitive behaviors exhibited by these mice.

Results

Cntnap2−/− dSPNs exhibit increased cortical drive

Emerging evidence indicates that corticostriatal synapses are a common site of alteration in mouse models of ASD (Li & Pozzo-Miller, 2020). To test whether mice with loss of Cntnap2 exhibit changes in corticostriatal connectivity, we crossed Cntnap2−/− mice to Thy1- ChR2-YFP mice, which express channelrhodopsin in a subset of layer V pyramidal neurons (Fig. 1A) (Arenkiel et al., 2007; Poliak et al., 1999). These mice were crossed to a D1-tdTomato reporter line to visually identify dSPNs (Ade et al., 2011). We recorded from SPNs in the dorsolateral striatum (DLS), as this sensorimotor striatal subregion is implicated in the acquisition of habitual and procedural behaviors (Packard & Knowlton, 2002). Changes in physiological function in this area may be connected to the acquisition of repetitive motor behaviors in ASD (APA, 2022; Fuccillo, 2016). To simulate a train of cortical inputs, we applied ten pulses of blue light over the recording site in DLS and measured the number of action potentials (APs) fired by SPNs in the absence of synaptic blockers (Fig. 1A). We altered the light intensity to vary the probability of eliciting subthreshold depolarizations or AP firing. dSPNs were identified using tdTomato fluorescence, and tdTomato negative neurons were designated putative iSPNs.

Cntnap2−/− dSPNs exhibit increased cortical drive.

(A) Left: schematic of the corticostriatal connectivity experiments. For corticostriatal excitability, cortical terminals expressing ChR2 were stimulated with 10 pulses of blue light at 10 Hz and responses were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. For corticostriatal synaptic strength, cortical terminals expressing ChR2 were stimulated with blue light at increasing intensity and synaptic currents were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. Center: 10x confocal image of the striatum from a Cntnap2+/+;D1-tdTomato;Thy1-ChR2 mouse. Right: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;D1-tdTomato;Thy1-ChR2 mouse. YFP (green) labels cell bodies and axons of a subset of layer V pyramidal neurons, tdTomato (red) labels dSPNs, and DAPI stained nuclei are in blue.

(B) Example single traces of action potentials (APs) in dSPNs evoked by cortical terminal stimulation at 20% light intensity for the indicated genotypes.

(C) Quantification (mean ± SEM) of the number of APs evoked in dSPNs at different light intensities. Cntnap2+/+ n=9 mice, 24 cells, Cntnap2−/− n=10 mice, 22 cells. Repeated measures two-way ANOVA p values are shown; g x s F (12, 204) = 1.935, geno F (1, 17) = 2.034, stim F (1.931, 32.83) = 86.12.

(D) Example single traces of action potentials (APs) in iSPNs evoked by cortical terminal stimulation at 20% light intensity for the indicated genotypes.

(E) Quantification (mean ± SEM) of the number of APs evoked in iSPNs at different light intensities. Cntnap2+/+ n=9 mice, 23 cells, Cntnap2−/− n=10 mice, 21 cells. Repeated measures two-way ANOVA p values are shown; g x s F (12, 216) = 0.5012, geno F (1, 18) = 0.9989, stim F (2.331, 41.96) = 60.62.

(F) Average EPSC traces from example dSPNs of each genotype induced by optogenetic cortical terminal stimulation at 14% light intensity.

(G) Quantification of EPSC amplitude evoked in dSPNs at different light intensities (line represents the mean, dots/squares represent average EPSC amplitude for each mouse). Cntnap2+/+ n=8 mice, 17 cells, Cntnap2−/− n=5 mice, 13 cells. Repeated measures two-way ANOVA p values are shown; g x s F (2, 22) = 0.2154, geno F (1, 11) = 0.2585, stim F (1.053, 11.58) = 49.68.

(H) Average EPSC traces from example iSPNs of each genotype induced by optogenetic cortical terminal stimulation at 14% light intensity.

(I) Quantification of EPSC amplitude evoked in iSPNs at different light intensities (line represents mean, dots/squares represent average for mouse). Cntnap2+/+ n=6 mice, 13 cells, Cntnap2−/− n=5 mice, 11 cells. Repeated measures two-way ANOVA p values are shown; g x s F (2, 18) = 0.4428, geno F (1, 9) = 0.5635, stim F (1.095, 9.851) = 23.82.

(J) Quantification (mean ± SEM) of AMPA:NMDA ratio in dSPNs evoked by 20% light intensity (dots/squares represent average AMPA:NMDA ratio for each mouse). Cntnap2+/+ n=5 mice, 22 cells, Cntnap2−/− n=5 mice, 22 cells, p=0.8413, Mann-Whitney test.

(K) Example traces show pairs of EPSCs evoked by optogenetic corticostriatal stimulation (20% light intensity) recorded at +40 mV (top traces) and −70 mV (bottom traces) from Cntnap2+/+ and Cntnap2−/− dSPNs.

(L) Quantification (mean ± SEM) of AMPA:NMDA ratio in iSPNs evoked by 20% light intensity (dots/squares represent average AMPA:NMDA ratio for each mouse). Cntnap2+/+ n=5 mice, 21 cells, Cntnap2−/− n=5 mice, 21 cells, p=0.3095, Mann-Whitney test.

(M) Example traces show pairs of EPSCs evoked by optogenetic corticostriatal stimulation (20% light intensity) recorded at +40 mV (top traces) and −70 mV (bottom traces) from Cntnap2+/+ and Cntnap2−/− iSPNs.

We quantified the number of evoked APs at different light intensities and found that dSPNs in young adult Cntnap2−/− mice exhibited increased spike probability compared to wild-type (WT) dSPNs (Fig. 1B-C). The interaction effect of genotype and stimulation intensity in these cells suggests increased corticostriatal drive, consistent with findings in another mouse model with loss of function of the ASD-risk gene Tsc1 (Benthall et al., 2021). Cntnap2−/− iSPNs had subtly increased cortically-evoked AP firing compared to WT iSPNs, although this was not statistically significant (1D-E). To test whether the enhanced spiking probability of Cntnap2−/− dSPNs was due to excitatory synaptic changes, we applied blue light of varying intensity over the recording site in DLS while holding cells at –70 mV to evoke AMPAR-driven excitatory postsynaptic currents (EPSCs). We found that the average optically-evoked EPSC amplitude was not significantly different across a range of light intensities in Cntnap2−/− dSPNs or iSPNs compared to WT (Fig. 1F-I). In an additional group of mice, we measured the ratio of AMPAR currents recorded at −70 mV to NMDAR currents recorded at +40 mV (at 20% blue light intensity) and found no significant differences in AMPA:NMDA ratio in Cntnap2−/− dSPNs or iSPNs (Fig. 1J-M).

To further assess synaptic inputs, we measured the number of dendritic spines in Cntnap2−/− and WT SPNs, which are typically the sites of cortical synaptic innervation (Bouyer et al., 1984; Xu et al., 1989). To visualize spines, we injected neonate Cntnap2;Drd1a-tdTomato mice with AAV5-Syn1-GFP virus to sparsely label dSPNs and iSPNs in the DLS (Fig. S1) (Keaveney et al., 2018). We found that Cntnap2−/− SPNs in adult mice had similar spine density as WT (Fig. S1), suggesting no overall change in synapse number. Together, these results show that dSPNs in Cntnap2−/− mice exhibit enhanced cortically-driven spiking. However, this is not due to a change in corticostriatal synaptic strength or overall synapse density.

Inhibition is unchanged in Cntnap2−/− SPNs

Previous work has indicated a reduction in the number and/or function of fast-spiking parvalbumin-expressing (PV) interneurons across multiple brain regions in Cntnap2−/− mice (Ahmed et al., 2023; Antoine et al., 2019; Jurgensen & Castillo, 2015; Paterno et al., 2021; Penagarikano et al., 2011; Vogt et al., 2018). While inhibitory deficits have been identified in the cortex and hippocampus (Antoine et al., 2019; Jurgensen & Castillo, 2015), and the number of PV interneurons has been reported to be decreased in striatum (Penagarikano et al., 2011), a comprehensive assessment of inhibitory synaptic function has yet to be completed in the striatum of Cntnap2−/− mice. We first determined whether there were global deficits in inhibition onto SPNs in Cntnap2−/− mice using intrastriatal electrical stimulation to evoke inhibitory postsynaptic currents (IPSCs) (Fig. 2A). In Cntnap2−/− dSPNs and iSPNs, the average amplitude of electrically-evoked IPSCs across a range of stimulation intensities was not different from WT (Fig. 2B-E).

Inhibition is not altered in Cntnap2−/− SPNs.

(A) Top: schematic of the experiment. A bipolar stimulating electrode was placed approximately 200 μm from the recording site. A range of electrical stimulation intensities was applied to the tissue while IPSCs were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. Bottom: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;D1-tdTomato mouse. tdTomato (red) labels dSPNs, DAPI stained nuclei are in blue.

(B) Average IPSC traces from example dSPNs of each genotype evoked by electrical stimulation at 1.5 (x0.01 mA) intensity for the indicated genotypes.

(C) Quantification (mean ± SEM) of IPSC amplitude in dSPNs at different stimulation intensities. Cntnap2+/+ n=17 cells from 9 mice, Cntnap2−/− n=16 cells from 9 mice. Repeated measures two-way ANOVA p values are shown; g x s F (7, 217) = 1.080, geno F (1, 31) = 0.0007751, stim F (1.815, 56.28) = 54.92.

(D) Average IPSC traces from example iSPNs of each genotype evoked by electrical stimulation at 1.5 (x0.01 mA) intensity for the indicated genotypes.

(E) Quantification (mean ± SEM) of IPSC amplitude in iSPNs at different stimulation intensities. Cntnap2+/+ n=16 cells from 9 mice, Cntnap2−/− n=16 cells from 10 mice. Repeated measures two-way ANOVA p values are shown; g x s F (7, 210) = 0.8741, geno F (1, 30) = 1.746, stim F (1.591, 47.73) = 45.66.

(F) Top: schematic of the experiment. PV interneuron terminals expressing ChR2 were stimulated with blue light at a range of intensities and optically-evoked IPSCs were recorded from dSPNs (red) and iSPNs (grey) in dorsolateral striatum. Bottom: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;D1-tdTomato;PV-Cre;Ai32 mouse. YFP (green) labels PV interneurons, tdTomato (red) labels dSPNs, DAPI stained nuclei are in blue.

(G) Average IPSC traces from example dSPNs of each genotype evoked by optogenetic PV interneuron stimulation at 30% light intensity.

(H) Quantification (mean ± SEM) of IPSC amplitude in dSPNs at different light intensities. Cntnap2+/+ n=29 cells from 15 mice, Cntnap2−/− n=23 cells from 11 mice. Repeated measures two-way ANOVA p values are shown; g x s F (7, 441) = 0.2566, geno F (1, 63) = 0.001322, stim F (1.433, 90.25) = 32.57.

(I) Average IPSC traces from example iSPNs of each genotype evoked by optogenetic PV interneuron stimulation at 30% light intensity.

(J) Quantification (mean ± SEM) of IPSC amplitude in iSPNs at different light intensities. Cntnap2+/+ n=24 cells from 14 mice, Cntnap2−/− n=27 cells from 13 mice. Repeated measures two-way ANOVA p values are shown; g x s F (7, 343) = 1.441, geno F (1, 49) = 0.4771, stim F (1.622, 79.46) = 38.49.

There are many sources of inhibition in the striatum (Burke et al., 2017), which can all be activated with electrical stimulation. To assess whether inhibition from PV interneurons specifically is altered in Cntnap2−/− mice, we crossed Cntnap2−/−;D1-tdTomato mice to Pvalb-Cre; RCL-ChR2-H134R-EYFP (Ai32) mice to express channelrhodopsin in PV interneurons (Fig. 2F) (Hippenmeyer et al., 2005; Madisen et al., 2010). We applied a blue light pulse of varying intensity over the recording site to evoke PV interneuron-specific IPSCs in SPNs, in the presence of excitatory synaptic blockers (Fig. 2F). We found that the average amplitude of optically-evoked IPSCs did not differ significantly in Cntnap2−/− dSPNs or iSPNs compared to WT controls (Fig. 2G-J).

To directly measure PV neuron function, we assessed the intrinsic excitability of PV interneurons in Cntnap2−/− mice. To visualize PV interneurons for recordings, we crossed Cntnap2−/− mice to Pvalb-Cre;RCL-tdT (Ai9) mice (Fig. S2A). Plotting the number of APs fired as a function of current step amplitude indicated that there were no significant differences in the intrinsic excitability of PV interneurons in Cntnap2−/− mice compared to controls (Fig. S2B,C). There were also no changes in intrinsic cell properties such as rheobase, membrane resistance, capacitance, resting membrane potential, or AP shape in Cntnap2−/− PV interneurons (Fig. S2D-K).

Given prior reports of altered PV cell number in Cntnap2−/− mice (Paterno et al., 2021; Penagarikano et al., 2011; Vogt et al., 2018), we counted PV-expressing cells in the striatum, using immunohistochemistry and fluorescent in situ hybridization. We found no significant difference in the number of PV-positive cells in the dorsal striatum of Cntnap2−/− mice compared to WT (Fig. S3A-F and K-N). We also found no changes in the average single cell or total level of PV protein expression in the dorsal striatum (Fig. 3G-J). Overall, we did not observe significant changes in PV interneuron number, PV expression, or PV interneuron-mediated inhibition in the adult Cntnap2−/− striatum compared to WT controls.

Intrinsic excitability is increased in Cntnap2−/− dSPNs.

(A) Example AP traces in dSPNs evoked by a 200 pA current step for the indicated genotypes.

(B) Quantification (mean ± SEM) of the number of APs evoked in dSPNs at different current step amplitudes. Cntnap2+/+ n=22 cells from 8 mice, Cntnap2−/− n=23 cells from 8 mice. Repeated measures two-way ANOVA p values are shown; g x c F (12, 528) = 2.649, geno F (1, 44) = 107.5, current F (1.974, 86.86) = 147.5.

(C) Quantification (mean ± SEM) of the rheobase current in dSPNs. Dots/squares represent the rheobase current for each neuron. n is the same as in panel B. **p=0.0016, two-tailed unpaired t test.

(D-F) Quantification (mean ± SEM) of dSPN membrane resistance (D), *p=0.0328, Mann-Whitney test; membrane capacitance (E), p=0.2182, Mann-Whitney test; and resting membrane potential (F), p=0.9914, two-tailed unpaired t test. Dots/squares represent the average value for each neuron. n is the same as in panel B.

(G) Example AP traces in iSPNs evoked by a 200 pA current step for the indicated genotypes.

(H) Quantification (mean ± SEM) of the number of APs evoked in iSPNs at different current step amplitudes. Cntnap2+/+ n=22 cells from 8 mice, Cntnap2−/− n=21 cells from 8 mice. Repeated measures two-way ANOVA p values are shown; g x c F (12, 516) = 1.569, geno F (1, 43) = 1.042, current F (2.041, 87.78) = 284.7.

(I) Quantification (mean ± SEM) of the rheobase current in iSPNs. Dots/squares represent the rheobase current for each neuron. n is the same as in panel H. p=0.0923, two-tailed unpaired t test.

(J-L) Quantification (mean ± SEM) of iSPN membrane resistance (J), p=0.8193, Mann-Whitney test; membrane capacitance (K), p=0.6886, two-tailed unpaired t test; and resting membrane potential (L), p=0.4859, two-tailed unpaired t test. Dots/squares represent the average value for each neuron. n is the same as in panel H.

dSPN intrinsic excitability is increased in Cntnap2−/− mice

Given that the increased cortical drive onto Cntnap2−/− dSPNs could not be explained by changes in excitatory or inhibitory synaptic function, we tested whether it could be due to a change in intrinsic excitability. To measure this, we recorded from dSPNs and iSPNs in Cntnap2;D1-tdTomato mice and injected current steps of increasing amplitude. We found that Cntnap2−/− dSPNs had significantly increased intrinsic excitability compared to WT dSPNs (Fig. 3A-B). Cntnap2−/− dSPNs also had reduced rheobase current (Fig. 3C), the minimum current required to evoke an AP, as well as increased membrane resistance (Fig. 3D). While there was a trend towards increased excitability in Cntnap2−/− iSPNs, this effect was not statistically significant (Fig. 3G-H), and these cells did not exhibit changes in rheobase current (Fig. 3I) or membrane resistance (Fig. 3J). Membrane capacitance (Fig. 3E,K), resting membrane potential (Fig. 3F,L), and AP shape (Fig. S4) were not significantly changed in Cntnap2−/− SPNs, although latency to first spike was reduced in both Cntnap2−/− dSPNs and iSPNs (Fig. S4C, I). Given the lack of synaptic changes observed in Cntnap2−/− dSPNs, the increase in dSPN intrinsic excitability likely underlies their enhanced corticostriatal drive (see Fig. 1).

The effects of Kv1.2 blockade are occluded in Cntnap2−/− dSPNs

Caspr2 is known to be involved in the clustering of voltage-gated potassium channels (Inda et al., 2006; Poliak et al., 1999; Poliak et al., 2003), particularly Kv1.2 channels (Scott et al., 2019). These channels play an important role in regulating the intrinsic excitability of SPNs (Nisenbaum et al., 1994). Blockade of these channels with the drug α-Dendrotoxin (α-DTX), results in increased AP frequency, decreased rheobase current, decreased first AP latency, and decreased AP threshold (Shen et al., 2004), particularly in dSPNs (Lahiri & Bevan, 2020). Given that Cntnap2−/− dSPNs exhibited increased AP frequency (Fig. 3B), decreased rheobase current (Fig. 3C), decreased first AP latency (Fig. S4C), and a trend towards decreased AP threshold (p=0.0516; Fig. S4B), we hypothesized that loss of function of Kv1.2 channels could be the mechanism. To test this, we recorded from dSPNs and iSPNs in Cntnap2;Drd1a-tdTomato mice and injected current steps of increasing amplitude in the absence or presence of α-DTX (100nM). We found that α-DTX enhanced AP firing and decreased AP threshold in WT dSPNs (Fig. 4A-C), but not in Cntnap2−/− dSPNs (Fig. 4D-F). AP width was also significantly increased by α-DTX in WT but not Cntnap2−/− dSPNs (Fig. S5H,Q), while AP fast afterhyperpolarization and AP latency were significantly decreased in both WT and Cntnap2−/− dSPNs (Fig. S5E, I, N, R). Together these data show that the effects of α-DTX on Cntnap2−/− dSPNs are largely occluded, indicating that Kv1.2 channels are basally altered by loss of Cntnap2. This likely accounts for the enhanced intrinsic excitability of Cntnap2−/− dSPNs.

The effects of Kv1.2 blockade are occluded in Cntnap2−/− dSPNs.

(A) Example AP traces from Cntnap2+/+ dSPNs evoked by a 200 pA current step in the absence or presence of α-Dendrotoxin (α-DTX).

(B) Quantification (mean ± SEM) of the number of APs evoked in Cntnap2+/+ dSPNs in the absence or presence of α-DTX at different current step amplitudes. Cntnap2+/+ n=17 cells from 8 mice, Cntnap2+/+ + α-DTX n=21 cells from 8 mice. Repeated measures two-way ANOVA p values are shown; drug x c F (13, 468) = 3.097, drug F (1, 36) = 2.399, current F (2.091, 75.28) = 172.0.

(C) Quantification (mean ± SEM) of the AP threshold in Cntnap2+/+ dSPNs in the absence or presence of α-DTX. Dots/triangles represent the threshold for each neuron. n is the same as in panel B. **p=0.0010, Mann-Whitney test.

(D) Example AP traces from Cntnap2−/− dSPNs evoked by a 200 pA current step in the absence or presence of α-DTX.

(E) Quantification (mean ± SEM) of the number of APs evoked in Cntnap2−/− dSPNs. Cntnap2−/− n=21 cells from 9 mice, Cntnap2−/− + α-DTX n=20 cells from 9 mice. Repeated measures two-way ANOVA p values are shown; drug x c F (13, 507) = 0.6054, drug F (1, 39) = 0.03352, current F (2.156, 84.07) = 247.8.

(F) Quantification (mean ± SEM) of the AP threshold in Cntnap2−/− dSPNs. Squares/triangles represent the threshold for each neuron. n is the same as in panel E. p=0.1348, Mann-Whitney test.

(G) Example AP traces in Cntnap2+/+ iSPNs evoked by a 200 pA current step in the absence or presence of α-DTX.

(H) Quantification (mean ± SEM) of the number of APs evoked in Cntnap2+/+ iSPNs. Cntnap2+/+ n=17 cells from 9 mice, Cntnap2+/+ + α-DTX n=17 cells from 9 mice. Repeated measures two-way ANOVA p values are shown; drug x c F (13, 416) = 0.3719, drug F (1, 32) = 2.161, current F (1.955, 62.57) = 215.9.

(I) Quantification (mean ± SEM) of the AP threshold in Cntnap2+/+ iSPNs. Dots/triangles represent the threshold for each neuron. n is the same as in panel H. p=0.5401, Mann-Whitney test.

(J) Example AP traces in Cntnap2−/− iSPNs evoked by a 200 pA current step in the absence or presence of α-DTX.

(K) Quantification (mean ± SEM) of the number of APs evoked in Cntnap2−/− iSPNs. Cntnap2−/− n=18 cells from 10 mice, Cntnap2−/− + α-DTX n=19 cells from 10 mice. Repeated measures two-way ANOVA p values are shown; drug x c F (13, 455) = 2.623, drug F (1, 35) = 0.01734, current F (1.721, 60.23) = 227.4.

(L) Quantification (mean ± SEM) of the AP threshold in Cntnap2−/− iSPNs. Squares/triangles represent the threshold for each neuron. n is the same as in panel K. p=0.4250, two-tailed unpaired t test.

We tested the effects of α-DTX on WT and Cntnap2−/− iSPNs and found minimal changes (Fig. 4G-K, Fig. S6), with no alterations in AP threshold (Fig. 4I,L) and only a small shift in the input-output relationship of Cntnap2−/− iSPNs (Fig. 4K). The effects of α-DTX on iSPN excitability have been less well documented and our data suggest a difference in the sensitivity of dSPNs and iSPNs to Kv1 blockade.

Cntnap2−/− mice display increased repetitive behaviors

RRBs comprise one of the primary symptom domains of ASD (APA, 2022). Alterations in striatal circuits are thought to be involved in the manifestation of RRBs, given the striatum’s role in action selection and motor control (Estes et al., 2011; Fuccillo, 2016; Hollander et al., 2005; Langen et al., 2014). To determine whether changes in motor behavior accompanied the altered striatal physiology in Cntnap2−/− mice, we assessed locomotor activity and spontaneous repetitive behaviors using the open field, marble burying, and holeboard assays (Fig. 5A-D). In the open field, we found no significant difference in the total distance traveled, average speed, or number of rears in Cntnap2−/− mice compared to WT controls (Fig. 5E-G). We did find that Cntnap2−/− mice made significantly more entries into the center of the open field arena than WT mice, which may reflect a reduction in avoidance behavior in these mice (Fig. 5H). Manually scored grooming behavior revealed that Cntnap2−/− mice initiated more grooming bouts than WT controls in the open field (Fig. 5I), consistent with prior reports (Penagarikano et al., 2011).

Cntnap2−/− mice have increased repetitive behaviors.

(A-D) Schematics of the behavioral assays used to measure repetitive behaviors in Cntnap2+/+ and Cntnap2−/− mice, created with BioRender.com. Male and female mice were used for all tests.

(E-H) Quantification (mean ± SEM) of open field activity over 60 minutes. (E) total distance traveled, p= 0.3538, Mann-Whitney test; (F) average speed, p=0.3832, Mann-Whitney test; (G) number of rears, p=0.1892, two-tailed unpaired t test; (H) number of center entries, *p=0.0101, two-tailed unpaired t test. Cntnap2+/+ n=41 mice, Cntnap2−/− n=34 mice.

(I) Quantification (mean ± SEM) of the number of manually scored grooming bouts in the first 20 minutes of the open field test, **p=0.0034, Mann-Whitney test, Cntnap2+/+ n=41 mice, Cntnap2−/− n=34 mice.

(J) Quantification (mean ± SEM) of total marbles buried in the marble burying assay. Cntnap2+/+ n=33 mice and Cntnap2−/− n=33 mice, *p=0.0396, two-tailed unpaired t test.

(K-M) Quantification (mean ± SEM) of performance in the holeboard assay. (K) Total number of nose pokes in 10 minutes, *p=0.0212, Mann-Whitney test; (L) nose pokes in the first five minutes, p=0.4811, two-tailed unpaired t test; and (M) nose pokes in the last five minutes, *p=0.0116, two-tailed unpaired t test. Cntnap2+/+ n=25 mice, Cntnap2−/− n=22 mice.

(N) Quantification (mean ± SEM) of the frequency of movement syllables (top 25 most frequent syllables) in the open field assay defined by Keypoint-MoSeq. Cntnap2+/+ n=13 mice and Cntnap2−/− n=11 mice, **p=0.0013 for grooming 6, **p=0.0013 for grooming 7, Kruskal-Wallis test with Dunn’s correction for multiple comparisons.

(O) Quantification (mean ± SEM) of the entropy of syllable transitions in the open field assay. Cntnap2+/+ n=13 mice and Cntnap2−/− n=11 mice, *p=0.0236, two-tailed unpaired t test.

(P) Accuracy of a Random Forest decoder trained on DeepLabCut basic locomotor data (Movement), Keypoint-MoSeq syllable usage data (Syllable usage), or Keypoint-MoSeq syllable transition data (Syllable transition) in distinguishing between Cntnap2+/+ and Cntnap2−/− mice. Dotted line represents chance performance.

For panels E-M and O, dots/squares represent the value for each mouse.

To further assess motor behaviors in Cntnap2−/− mice, we utilized the marble burying assay (Fig. 5C), which takes advantage of a mouse’s natural tendency to dig or bury. The number of marbles buried is used as a measure of persistent or repetitive behavior (Angoa-Perez et al., 2013). We found that Cntnap2−/− mice buried significantly more marbles on average than WT controls (Fig. 5J). Another measure of repetitive behavior, which is based on the natural exploratory behavior of mice is the holeboard assay (Fig. 5D). In this test, the number of nose pokes made into unbaited holes is recorded. Cntnap2−/− mice made significantly more nose pokes within a 10-minute period than WT mice (Fig. 5K). This was largely due to increased poking during the last 5 minutes of the test (Fig. 5L-M), indicating persistent poking behavior in Cntnap2−/− mice. Together, the increased grooming, marble burying and nose poking indicate an increase in RRBs in Cntnap2−/− mice. A summary of behavior test results by genotype and sex is shown in Table S1.

To gain further insight into the spontaneous behavior profile of Cntnap2−/− mice, we utilized a combination of DeepLabCut and Keypoint-MoSeq to perform unbiased, machine learning-based assessment of general locomotion and behavior in an additional cohort of Cntnap2−/− mice (Fig. 5N-P, Fig. S7) (Mathis et al., 2018; Wiltschko et al., 2020). Again, we found that Cntnap2−/− mice did not exhibit changes in general locomotor activity compared to WT littermates (Fig. S7A). Analysis of movement “syllables” using Keypoint-MoSeq revealed that across the 25 most frequent syllables, two syllables associated with grooming were performed with significantly increased frequency in Cntnap2−/− mice (Fig. 5N). Cntnap2−/− mice also had an increase in the total number of grooming bouts (Fig. S7B), replicating the findings in the manually scored cohort (see Fig. 5I). While syllable usage was generally similar between WT and Cntnap2−/− mice, transitions between syllables differed between the groups (Fig. S7C). A measure of the entropy of syllable transitions revealed that Cntnap2−/− mice exhibited less entropy, suggesting less variability in the transition from one movement syllable to the next (Fig. 5O). This rigidity in motor sequence may be indicative of more restricted motor behavior overall. Finally, we tested whether a trained decoder could accurately distinguish WT and Cntnap2−/− mice using information about movement, syllable usage, or syllable transitions. The decoding models performed significantly better than chance at identifying WT and Cntnap2−/− mice based on their syllable usage and transitions, but not general locomotor activity (Fig. 5P). Together, this analysis demonstrates that while overall locomotor activity is not strongly affected in Cntnap2−/− mice, the behavior patterns of these mice are distinct from WT, reflecting enhanced repetitive behaviors.

Cntnap2−/− mice exhibit enhanced motor learning

The accelerating rotarod is a striatal-dependent measure of motor coordination and learning that has been used across a range of ASD mouse models (Cording & Bateup, 2023). Changes in corticostriatal circuits have been identified in mouse models of ASD with altered performance in the task (Cording & Bateup, 2023). Given the altered corticostriatal drive in Cntnap2−/− mice, we tested whether motor coordination and learning were affected in these mice. In the rotarod test, mice learn to walk and then run to stay on a rotating rod as it increases in speed over the course of five minutes. Mice perform three trials a day for four days. In trials one through six, the rod increases in speed from five to 40 revolutions per minute (RPM), while in trials seven through 12 the rod increases from 10 to 80 RPM (Fig. 6A). Learning occurs over trials within a day, as well as across days, as the mouse develops and hones a stereotyped motor pattern to stay on the rod for increasing amounts of time (Rothwell et al., 2014; Yin et al., 2009). We found that Cntnap2−/− mice performed significantly better than WT mice in this task, particularly in the later trials when the rod rotates at the faster 10 to 80 RPM speed (Fig. 6B).

Cntnap2−/− mice exhibit enhanced motor learning.

(A) Schematic of the rotarod apparatus (top), and design of the task (bottom), created with BioRender.com. Mice walk to stay on the rotating rod for three 5-minute trials a day for two days at 5-40 RPM acceleration over 5 minutes, followed by three trials a day for two days at 10-80 RPM.

(B) Quantification (mean ± SEM) of accelerating rotarod performance across 12 trials for the indicated genotypes. Cntnap2+/+ n=30 mice, Cntnap2−/− n=29 mice. Repeated measures two-way ANOVA p values are shown; g x t F (11, 616) = 4.935, geno F (1, 56) = 15.29, trial F (7.245, 405.7) = 108.4.

(C) Quantification (mean ± SEM) of rotarod performance on trial 1 quantified as terminal speed. Dots/squares represent the performance of individual mice. n is same as in panel B, p=0.1518, Mann-Whitney test.

(D) Quantification (mean ± SEM) of learning rate (RPM/day) calculated as the slope of the line of performance from the first trial (1) to the last trial (12) for each mouse. Dots/squares represent the learning rate for individual mice. n is same as in panel B, ***p=0.0002, two-tailed unpaired t test.

Initial performance (terminal velocity on trial one) was not significantly different between WT and Cntnap2−/− mice (Fig. 6C), but the learning rate from trial one to trial 12 was increased in Cntnap2−/− mice (Fig. 6D). These findings expand upon previous work indicating increased performance on both steady-state and accelerating rotarod tasks utilizing slower speeds in Cntnap2−/− mice (Dawes et al., 2018; Penagarikano et al., 2011). These results also align with the increased rotarod performance seen in other ASD mouse models exhibiting enhanced corticostriatal drive (Benthall et al., 2021; Cording & Bateup, 2023).

Cntnap2−/− mice exhibit cognitive inflexibility

RRBs include not just stereotyped movements, but also insistence on sameness and perseverative interests (APA, 2022). Cognitive inflexibility, a deficit in the ability to flexibly adapt to changes in the environment and update behavior, is a manifestation of ASD and some other psychiatric disorders, which are associated with striatal dysfunction (Fuccillo, 2016). Indeed, in individuals with ASD, the severity of RRBs is associated with measures of cognitive inflexibility, and evidence from imaging studies suggests that altered corticostriatal connectivity may be present in the case of both repetitive behaviors and cognitive inflexibility (Uddin, 2021). To assess cognitive flexibility in Cntnap2−/− mice, we utilized a four-choice odor-based reversal learning assay (Johnson et al., 2016; Lin et al., 2022). Briefly, mice were trained to dig for a reward in one of four pots containing scented wood shavings (Fig. 7A). On the first day of the task (acquisition), the rewarded pot was scented with odor one (O1). Mice reached criterion when they chose O1 for at least eight of 10 consecutive trials. On day two, mice were given a recall test in which the rewarded pot was again scented with O1. After reaching criterion, the reversal trials began, and the rewarded pot was scented with the previously unrewarded odor two (O2). To reach criterion, mice must learn the new association of O2 and reward and choose O2 for eight of 10 consecutive trials.

Cntnap2−/− mice demonstrate cognitive inflexibility.

(A) Schematic of the four-choice odor-based reversal learning task, created with BioRender.com. Colored circles represent pots with different scented wood shavings. Yellow ring represents the food reward. Red circle in the Day 2 Reversal panel indicates a novel odor.

(B-D) Quantification of parameters during acquisition learning. Mean ± SEM number of trials to reach criterion (B, at least 8 out of last 10 trials correct), p=0.5397, Mann-Whitney test; number of quadrant entries before making a choice (C), p=0.9118, Mann-Whitney test; and latency to make a choice (D), p=0.7224, two-tailed unpaired t test. Dots/squares represent the value for each mouse.

(E) Quantification (mean ± SEM) of the number of trials to reach criterion (at least 8 out of last 10 trials correct) during the recall test on day 2, p=0.3737, Mann-Whitney test.

(F-I) Quantification of parameters during reversal learning. Mean ± SEM number of trials to reach criterion (F, at least 8 out of last 10 trials correct), **p=0.0048, two-tailed unpaired t test; number of quadrant entries before making a choice (G), *p=0.0158, two-tailed unpaired t test; latency to make a choice (H), **p=0.0013, two-tailed unpaired t test; and number of trials to reach criterion after the first correct choice (I), *p=0.0183, two-tailed unpaired t test.

(J) Quantification (mean ± SEM) of the total number of errors made during reversal learning, **p=0.0034, Mann-Whitney test.

(K) Quantification (mean ± SEM) of the different error types made during reversal learning. Perseverative errors, ***p=0.0005, regressive errors **p=0.0068, novel errors, p=0.9955, irrelevant errors p=0.9988, omissions, p=>0.9999, repeated measures two-way ANOVA with Šídák’s multiple comparisons test; g x e F (4, 72) = 4.292, geno F (1, 18) = 10.91, error F (4, 72) = 53.49.

(L) Quantification (mean ± SEM) of the percent of choices made for each odor, binned across three trials, during reversal learning. Odor 1 was rewarded during acquisition learning. Odor 2 was rewarded during reversal learning. Odor 3 was never rewarded (irrelevant). Odor 4 was a novel odor introduced during the reversal learning phase.

For panels B-L, n=10 Cntnap2+/+ mice and 10 Cntnap2−/− mice.

During acquisition, Cntnap2−/− mice performed similarly to WT controls, not differing in the average number of trials needed to reach criterion, the number of quadrant entries made before making a choice, or the latency to choose a pot (Fig. 7B-D). On day two, Cntnap2−/− mice performed similarly to controls during recall, demonstrating successful consolidation of the odor-reward pairing (Fig. 7E). However, we found that Cntnap2−/− mice had a deficit in reversal learning, requiring significantly more trials on average than WTs to reach criterion once the odor-reward pairing was changed (Fig. 7F). Interestingly, during reversal, Cntnap2−/− mice made fewer quadrant entries before making a digging choice and had significantly decreased latency to make a choice compared to controls (Fig. 7G,H). Even after the first correct choice of O2 during reversal, Cntnap2−/− mice took more trials to reach criterion than WTs (Fig. 7I). In terms of errors, Cntnap2−/− mice made more reversal errors than WT mice (Fig. 7J), in particular perseverative (continuing to choose O1) and regressive (choosing O1 after correctly choosing O2 once) errors (Fig. 7K). Cntnap2−/− mice did not differ from WT controls in choices of the novel (newly introduced during reversal) or irrelevant (never rewarded) odors, or in the number of omitted trials (timing out without making a choice) (Fig. 7K). Instead, the persistence in choosing O1, even after at least one correct choice of O2, drove the cognitive inflexibility in these mice (Fig. 7L). This persistence in choice may be reflective of the broader scope of RRBs in Cntnap2−/− mice.

Discussion

In this study we tested whether loss of the ASD risk gene Cntnap2 altered striatal physiology or striatal-dependent behaviors. We found that direct pathway SPNs exhibited enhanced cortical drive in Cntnap2−/− mice. This change was not due to differences in excitatory or inhibitory synapses, as cortical inputs onto SPNs were unchanged and there were no significant deficits in inhibition onto SPNs in these mice. Instead, loss of Cntnap2 resulted in a significant increase in the excitability of dSPNs, likely driven by altered contribution of Kv1.2 channels to intrinsic firing properties. At the behavioral level, Cntnap2−/− mice exhibited repetitive behaviors including increased grooming, nose poking, and marble burying. These mice also had enhanced motor learning, performing significantly better than controls in the accelerating rotarod task. Finally, Cntnap2−/− mice exhibited cognitive inflexibility in the four-choice reversal learning assay.

Cellular phenotypes of Cntnap2 loss

The loss of Caspr2 has a variable impact on intrinsic excitability across brain regions and cell types. The increased intrinsic excitability that we identified in dSPNs has also been observed in Purkinje cells of the cerebellum (Fernandez et al., 2021) and pyramidal cells of the cortex (Antoine et al., 2019; Cifuentes-Diaz et al., 2023). However, we note hypoactivity (Brumback et al., 2018) or unchanged excitability (Lazaro et al., 2019) of pyramidal cells in some cortical regions in Cntnap2−/− mice. Caspr2 is involved in the clustering of voltage-gated potassium channels, in particular at the juxtaparanodes of myelinated axons (Poliak et al., 1999; Poliak et al., 2003) and the axon initial segment (Inda et al., 2006). Indeed, there are profound deficits in the clustering of Kv1-family channels in Cntnap2−/− mice, particularly Kv1.2 channels (Scott et al., 2019). These channels play an important role in regulating the intrinsic excitability of SPNs (Nisenbaum et al., 1994), in particular dSPNs (Lahiri & Bevan, 2020), and when blocked, result in increased excitability (Shen et al., 2004). The loss of Caspr2 in Cntnap2−/− mice may result in improper localization of Kv1.2 channels, and thus alter their contribution to intrinsic excitability. Indeed, we found that the effects of Kv1.2 blockade were largely occluded in Cntnap2−/− dSPNs, indicating that Kv1.2 loss of function is likely the mechanism driving the change in excitability. Interestingly, we find that blocking Kv1.2 channels has less of an effect on the excitability of iSPNs, which may account for the greater impact of Cntnap2 loss on dSPNs.

Prior studies of Cntnap2−/− mice have identified changes in the number of PV-expressing interneurons in the cortex (Penagarikano et al., 2011; Vogt et al., 2018), hippocampus (Paterno et al., 2021; Penagarikano et al., 2011) and striatum (Penagarikano et al., 2011). However, this finding is inconsistent across studies, as others have reported no change in the number of PV interneurons in these regions (Ahmed et al., 2023; Lauber et al., 2018; Scott et al., 2019). One possible explanation for this disparity is altered PV protein expression in Cntnap2−/− mice such that immunoreactivity varies in cell counting assessments. This is supported by the finding that the number of Vicia Villosa Agglutinin-positive (VVA+) perineuronal nets that preferentially surround PV cells is unchanged in Cntnap2−/− mice, even when PV immunoreactivity varies (Hartig et al., 1992; Haunso et al., 2000; Lauber et al., 2018). Parvalbumin, a Ca2+ buffer, plays an important role in the intrinsic fast-spiking properties of PV interneurons, such that a reduction in PV protein expression is known to change PV intrinsic function (Orduz et al., 2013). However, altered intrinsic properties of PV interneurons has also been variably reported across brain regions and studies of Cntnap2−/− mice, with subtle changes in PV firing properties reported in the developing striatum (Ahmed et al., 2023) and adult cortex (Vogt et al., 2018), but unchanged in the hippocampus (Paterno et al., 2021) and medial prefrontal cortex (Lazaro et al., 2019). In this study, we find no significant change in the number of PV interneurons or the striatal expression of PV protein in Cntnap2−/− mice. Consistent with this, we find no deficits in PV-mediated inhibition onto SPNs. Together, this suggests that primary changes in PV interneurons are unlikely to account for altered striatal circuit function in Cntnap2−/− mice.

Loss of Cntnap2 alters striatal-dependent behaviors

The striatum can be separated into functionally distinct subregions. We focused on the dorsal striatum in this study because of its role in controlling motor and cognitive functions (Voorn et al., 2004), which are relevant to ASD (Fuccillo, 2016; Subramanian et al., 2017). The dorsal striatum can be further subdivided into the dorsomedial striatum (DMS) and the dorsolateral striatum (DLS), with the former considered an associative region involved in goal-directed action-outcome learning and the latter implicated in the acquisition of habitual or procedural behaviors (Packard & Knowlton, 2002). We focused on cellular properties in the DLS as stereotyped, perseverative or persistent behaviors likely recruit DLS circuitry (Evans et al., 2024; Fuccillo, 2016). In the accelerating rotarod assay, learning and performance in the task has been associated with changes in the DLS. Positive modulation of the firing rate of DLS neurons occurs during rotarod training, in particular in later trials of the task, and synaptic potentiation of DLS SPNs in late training is necessary for intact performance (Yin et al., 2009). In line with this, lesions of the DLS impair both early and late rotarod learning (Yin et al., 2009). We found that Cntnap2−/− mice had increased rotarod performance, most notably at the later stages when DLS function is strongly implicated. Functionally, we also found increased cortical drive of DLS dSPNs in these mice, a change that was sufficient to increase rotarod performance in another mouse model with disruption of an ASD-risk gene (Benthall et al., 2021). Together, this supports a connection between the change observed in DLS SPN physiology and the increased motor routine learning in Cntnap2−/− mice, although this idea remains to be causally tested.

In terms of restricted, repetitive behaviors, we replicated prior studies showing increased spontaneous grooming in Cntnap2−/− mice (Penagarikano et al., 2011). Early evidence implicates the striatum in the control of the syntax or sequence of movements in a rodent grooming bout, such that very small lesions of DLS are capable of disrupting grooming (Cromwell & Berridge, 1996). However, recent work has also outlined roles for cellular modulation in DMS and ventral striatal Islands of Calleja in the control of grooming behavior (Ramirez-Armenta et al., 2022; Zhang et al., 2021). Cntnap2−/− mice also exhibited increased marble burying and nose poking.

The precise neurobiological substrates of these behaviors are yet unclear, but evidence linking increases in these behaviors to changes in cortico-striatal and amygdala-striatal function supports the notion that these behaviors may fit into a broader basal ganglia-associated RRB-like domain (Albelda & Joel, 2012; Lee et al., 2024).

In the four-choice reversal learning task, Cntnap2−/− mice showed no differences during the acquisition phase, suggesting that there were no broad deficits in learning. However, in the reversal stage of the task, Cntnap2−/− mice took significantly more trials to learn a new odor-reward pairing, owing primarily to continued choice of the previously rewarded odor. The DMS and ventral striatum (nucleus accumbens) have been shown to play an important role in reversal learning (Izquierdo et al., 2017), and in the four choice task specifically (Delevich et al., 2022). Additionally, decreased dopamine release in the DLS is associated with deficits in reversal learning in this task (Kosillo et al., 2019; Lin et al., 2022). Together, the learning phenotypes seen in Cntnap2−/− mice in the accelerating rotarod and reversal learning assay share an underlying rigidity in behavioral choice. In both cases, changes in striatal circuits likely underlie the repetitive, stereotyped behaviors.

In summary, our results fit into a model whereby divergent cellular changes in the striatum driven by a functionally diverse set of ASD risk genes similarly enhance corticostriatal drive, in particular, of the direct pathway. This in turn may facilitate striatal-dependent motor routine learning and behavioral perseveration. We speculate that a shared gain-of-function in striatal circuits may play a role in the formation of perseverative or repetitive behaviors in a sub-set of ASDs more broadly.

Limitations and future directions

This study characterized multiple striatal cell types and synapses as well as striatum-associated behaviors in Cntnap2−/− mice for the first time. However, there remain open questions as to the impact that Cntnap2 loss has on striatal function. Although we did not identify excitatory or inhibitory synaptic changes in Cntnap2−/− SPNs in this study, we only focused on a subset of these connections. While cortical inputs are a major source of excitation onto SPNs, there are other excitatory inputs onto these cells, such as from thalamus, that were not assessed in this study (Ding et al., 2008; Doig et al., 2010; Gerfen & Surmeier, 2011). In addition, while all intrastriatal GABAergic interneurons would have been sampled in the electrical stimulation experiments in this study, it is possible that interrogation of individual interneuron subtypes would reveal changes in specific inhibitory connections. Finally, major modulators of SPN activity such as cholinergic interneurons and dopaminergic inputs were not assessed here. These inputs have been implicated in several ASD mouse models (Kosillo & Bateup, 2021; Paval & Miclutia, 2021; Rapanelli et al., 2017), and changes in striatal cholinergic interneuron function have been identified in young (P21) Cntnap2−/− mice, making further study of this circuit particularly cogent (Ahmed et al., 2023).

The major cellular phenotype we observed was enhanced intrinsic excitability of Cntnap2−/− dSPNs. Given that Kv1.2 channels are known to be organized in part by Caspr2 (Poliak et al., 1999; Poliak et al., 2003; Scott et al., 2019) and the fact that blockade of Kv1.2 did not affect the excitability of Cntnap2−/− dSPNs, we conclude that loss of Cntnap2 leads to the improper clustering, number, or function of Kv1.2. However, more direct measurement of the number and/or localization of these channels through imaging, or the function of these channels through voltage clamp measurement of potassium currents, would bolster this conclusion. Further, assessing whether similar occlusion of the effects of blocking Kv1.2 channels occurs in the cortical and cerebellar cell types that have also been shown to be hyperexcitable in Cntnap2−/− mice would support a more holistic understanding of the impact of Cntnap2 loss on neuronal function (Antoine et al., 2019; Cifuentes-Diaz et al., 2023; Fernandez et al., 2021). Considering that these cell types also exhibit α-DTX-sensitive currents that boost AP firing, it is possible that the mechanism by which Cntnap2 loss increases excitability may be shared (Guan et al., 2007; Haghdoust et al., 2007; Khavandgar et al., 2005).

Finally, while we did identify changes in several striatum-associated behaviors in Cntnap2−/− mice, a causative relationship between the physiological and behavioral changes that we identified has not been established. Given that hyperexcitability of the SPNs of the movement-initiating direct pathway is the primary physiological change we identified, testing whether this change is necessary (i.e. by decreasing dSPN activity in Cntnap2−/− mice through cell type-specific expression of inward rectifying potassium channel Kir2.1) or sufficient (i.e. by increasing dSPN activity in WT mice using cell type-specific Gq-coupled (hM2Dq) DREADD activation) to alter behavior could illuminate the relationship between striatal function and behavior in Cntnap2−/− mice.

Supplementary Figures and Table

Cntnap2−/− SPNs do not have altered spine density.

(A) Representative deconvoluted images of dendritic spines from dSPNs for the indicated genotypes.

(B) Quantification (mean ± SEM) of dendritic spine density per 10 μm of dendrite in Cntnap2+/+ and Cntnap2−/− dSPNs. Dots/squares represent the average spine density per neuron. Cntnap2+/+ n=9 neurons (15 dendrites) from 6 mice, Cntnap2−/− n=8 neurons (15 dendrites) from 6 mice. p=0.9964, two-tailed unpaired t test.

(C) Representative deconvoluted images of dendritic spines from iSPNs for the indicated genotypes.

(D) Quantification (mean ± SEM) of dendritic spine density per 10 μm of dendrite in Cntnap2+/+ and Cntnap2−/− iSPNs. Dots/squares represent the average spine density per neuron. Cntnap2+/+ n=7 neurons (15 dendrites) from 4 mice, Cntnap2−/− n=10 neurons (15 dendrites) from 7 mice. p=0.5362, Mann-Whitney test.

PV interneuron intrinsic excitability is unchanged in Cntnap2−/− mice.

(A) Top: schematic of the experiment. tdTomato-expressing PV interneurons were recorded in the dorsolateral striatum and injected with current steps of increasing magnitude to evoke firing. Bottom: 20x confocal image of dorsolateral striatum from a Cntnap2+/+;PV-Cre;Ai9 mouse. tdTomato (red) labels PV interneurons, DAPI stained nuclei are in blue.

(B) Example traces of APs in PV interneurons evoked by a 200 pA current step for the indicated genotypes.

(C) Quantification (mean ± SEM) of the number of APs evoked in PV interneurons at different current step amplitudes. Cntnap2+/+ n=28 cells from 6 mice, Cntnap2−/− n=23 cells from 4 mice. Repeated measures two-way ANOVA p values are shown; g x c F (40, 1960) = 0.1251, geno F (1, 49) = 0.06951, current F (2.033, 99.60) = 107.5.

(D) Quantification (mean ± SEM) of the rheobase current. Dots/squares represent the rheobase current for each neuron. n is the same as in panel C. p=0.8852, two-tailed unpaired t test.

(E-G) Quantification (mean ± SEM) of the membrane resistance (E), p=0.3422, Mann-Whitney test; membrane capacitance (F), p=0.9055, two-tailed unpaired t test; and resting membrane potential (G), p=0.3141, two-tailed unpaired t test. Dots/squares represent the average value for each neuron. n is the same as in panel C.

(H) Phase plane plot of a single AP evoked in PV interneurons from Cntnap2+/+ (black) and Cntnap2−/− (red) mice.

(I-K) Quantification (mean ± SEM) of the AP threshold (I), p=0.6025, two-tailed unpaired t test; AP height (J), p=0.2850, two-tailed unpaired t test; and AP width at half max (K), p=0.2561, two-tailed unpaired t test. Dots/squares represent the value for the first spike evoked at rheobase for each neuron. n is the same as in panel C.

Cntnap2−/− mice do not exhibit changes in PV-positive cell number or expression.

(A) Representative confocal images of dorsolateral striatum labeled with an antibody against Parvalbumin (PV) protein for the indicated genotypes, mice aged postnatal day (P) 21.

(B) Quantification (mean ± SEM) of the number of PV positive cells per ROI at P21, p=0.2554, two-tailed unpaired t test. Cntnap2+/+ n=8 sections from 4 mice (2 sections imaged per mouse), Cntnap2−/− n=8 sections from 4 mice (2 sections imaged per mouse).

(C) Representative confocal images of dorsolateral striatum labeled with an antibody against PV for the indicated genotypes, mice aged P40-50.

(D) Quantification (mean ± SEM) of the number of PV positive cells per ROI, p=0.5144, two-tailed unpaired t test. Cntnap2+/+ n=24 sections from 12 mice (2 sections imaged per mouse), Cntnap2−/− n=24 sections from 12 mice (2 sections imaged per mouse).

(E) Representative confocal images of dorsal striatum labeled with an antibody against PV for the indicated genotypes, mice aged P40-50.

(F) Quantification (mean ± SEM) of the number of PV positive cells in the whole dorsal striatum at P40-50, p=0.4201, two-tailed unpaired t test. Cntnap2+/+ n=8 sections from 4 mice (2 sections imaged per mouse), Cntnap2−/− n=8 sections from 4 mice (2 sections imaged per mouse).

(G) Quantification (mean ± SEM) of the average single cell fluorescence of PV antibody staining in dorsal striatum, p=0.9996, two-tailed unpaired t test. Cntnap2+/+ n=8 sections from 4 mice (2 sections imaged per mouse), Cntnap2−/− n=8 sections from 4 mice (2 sections imaged per mouse).

(H) Quantification (mean ± SEM) of the average bulk fluorescence of PV antibody staining in dorsal striatum per section, p=0.4878, two-tailed unpaired t test. Cntnap2+/+ n=8 sections from 4 mice (2 sections imaged per mouse), Cntnap2−/− n=8 sections from 4 mice (2 sections imaged per mouse).

(I) Representative western blots for CASPR2, PV, and GAPDH protein in dorsal striatal tissue punches. 10 independent samples per genotype are shown.

(J) Quantification (mean ± SEM) of PV protein expression normalized to GAPDH. Data are presented as a percentage of Cntnap2+/+ levels, p=0.1485, two-tailed unpaired t test. Cntnap2+/+ n=10 mice, Cntnap2−/− n=10 mice.

(K) Representative confocal images of in situ hybridization for Pvalb in dorsolateral striatum for the indicated genotypes, mice aged P40-50.

(L-N) Quantification (mean ± SEM) of the number of Pvalb positive cells in the dorsolateral striatum per section (L), p>0.9999, Mann-Whitney test; dorsomedial striatum (M), p=0.7429, Mann-Whitney test; and central striatum (N), p=0.6286, Mann-Whitney test. Cntnap2+/+ n=4 sections from 2 mice (2 sections imaged per mouse), Cntnap2−/− n=4 sections from 2 mice (2 sections imaged per mouse).

Latency to spike is reduced in Cntnap2−/− SPNs.

(A) Phase plane plot of a single AP evoked in dSPNs from Cntnap2+/+ (black) and Cntnap2−/− (orange) mice.

(B) Quantification (mean ± SEM) of the dSPN AP threshold. Dots/squares represent the threshold for each neuron. Cntnap2+/+ n=22 cells from 8 mice, Cntnap2−/− n=23 cells from 8 mice, p=0.0516, two-tailed unpaired t test.

(C) Quantification (mean ± SEM) of dSPN AP latency to first spike plotted by number of APs fired in the current pulse. Shaded area around line represents SEM. Mixed-effects model with Geisser-Greenhouse correction p values are shown; g x APs F (19, 387) = 7.870, geno F (1, 43) = 24.98, APs F (1.740, 35.44) = 913.5.

(D-F) Quantification (mean ± SEM) of the dSPN AP fast afterhyperpolarization (D), p=0.4664, two-tailed unpaired t test; AP height (E), p=0.0833, two-tailed unpaired t test; and AP width at half max (F), p=0.3792, two-tailed unpaired t test. Dots/squares represent the value for the first spike evoked at rheobase for each neuron, n same as in panel B.

(G) Phase plane plot of a single AP evoked in iSPNs in Cntnap2+/+ (green) and Cntnap2−/− (blue) mice.

(H) Quantification (mean ± SEM) of the iSPN AP threshold. Dots/squares represent the threshold for each neuron. Cntnap2+/+ n=22 cells from 8 mice, Cntnap2−/− n=21 cells from 8 mice, p=0.1495, two-tailed unpaired t test.

(I) Quantification (mean ± SEM) of iSPN AP latency to first spike plotted by number of APs fired in the current pulse. Shaded area around line represents SEM. Mixed-effects model with Geisser-Greenhouse correction p values are shown; g x APs F (19, 343) = 3.300, geno F (1, 41) = 45.55, APs F (1.817, 32.81) = 680.4.

(J-L) Quantification (mean ± SEM) of iSPN AP fast afterhyperpolarization (J), p=0.3281, two-tailed unpaired t test; AP height (K), p=0.1794, Mann-Whitney test; and AP width at half max (L), p=0.2409, two-tailed unpaired t test. Dots/squares represent the value for the first spike evoked at rheobase for each neuron, n same as in panel H.

Inhibition of Kv1.2 differentially impacts action potential properties in Cntnap2+/+ and Cntnap2−/− dSPNs.

(A) Phase plane plot of a single AP evoked in Cntnap2+/+ dSPNs in the absence (solid black) or presence (dotted black) of α-Dendrotoxin (α-DTX).

(B-C) Quantification (mean ± SEM) of the Cntnap2+/+ dSPN membrane resistance (B), p=0.2681, Mann-Whitney test; and membrane capacitance (C), p=0.9886, two-tailed unpaired t test. Dots/triangles represent the average value for each neuron, Cntnap2+/+ n=17 cells from 8 mice, Cntnap2+/+ + α-DTX n=21 cells from 8 mice.

(D) Quantification (mean ± SEM) of the rheobase current in Cntnap2+/+ dSPNs. Dots/triangles represent the rheobase current for each neuron. n is the same as in panels B-C. p=0.2703, two-tailed unpaired t test.

(E-H) Quantification (mean ± SEM) of the Cntnap2+/+ dSPN fast afterhyperpolarization (E), *p=0.0274, two-tailed unpaired t test; resting membrane potential (F), p=0.1207, two-tailed unpaired t test; AP height (G), p=0.3037, two-tailed unpaired t test; and AP width at half max (H), **p=0.0084, two-tailed unpaired t test. Dots/triangles represent the value for the first spike evoked at rheobase for each neuron. n is the same as in panels B-C.

(I) Quantification (mean ± SEM) of Cntnap2+/+ dSPN AP latency to first spike, plotted by number of APs fired in the current pulse. Shaded area around line represents SEM. Mixed-effects model with Geisser-Greenhouse correction p values are shown; d x APs F (19, 126) = 4.825, drug F (1, 36) = 23.33, APs F (19.00, 126.0) = 106.7.

(J) Phase plane plot of a single AP evoked in Cntnap2−/− dSPNs in the absence (solid orange) or presence (dotted orange) of α-Dendrotoxin (α-DTX).

(K-L) Quantification (mean ± SEM) of the Cntnap2−/− dSPNs membrane resistance (K), *p=0.0452, Mann-Whitney test; and membrane capacitance (L), p=0.4950, two-tailed unpaired t test. Squares/triangles represent the average value for each neuron, Cntnap2−/− n=21 cells from 9 mice, Cntnap2−/− + α-DTX n=20 cells from 9 mice.

(M) Quantification (mean ± SEM) of the rheobase current in Cntnap2−/− dSPNs. Squares/triangles represent the rheobase current for each neuron. n is the same as in panels K-L. p=0.6239, Mann-Whitney test.

(N-Q) Quantification (mean ± SEM) of the Cntnap2−/− dSPN fast afterhyperpolarization (N), **p=0.0017, two-tailed unpaired t test; resting membrane potential (O), p=0.5681, two-tailed unpaired t test; AP height (P), p=0.0652, two-tailed unpaired t test; and AP width at half max (Q), p=0.3400, two-tailed unpaired t test. Squares/triangles represent the value for the first spike evoked at rheobase for each neuron. n is the same as in panels K-L.

(R) Quantification (mean ± SEM) of Cntnap2−/− dSPN AP latency to first spike, plotted by number of APs fired in the current pulse. Shaded area around line represents SEM. Mixed-effects model with Geisser-Greenhouse correction p values are shown; d x APs F (19, 135) = 3.362, drug F (1, 39) = 30.66, APs F (19.00, 135.0) = 123.8.

Inhibition of Kv1.2 does not strongly affect the excitability of iSPNs.

(A) Phase plane plot of a single AP evoked in Cntnap2+/+ iSPNs in the absence (solid green) or presence (dotted green) of α-Dendrotoxin (α-DTX).

(B-C) Quantification (mean ± SEM) of the Cntnap2+/+ iSPN membrane resistance (B), p=0.6624, two-tailed unpaired t test; and membrane capacitance (C), p=0.6444, two-tailed unpaired t test. Dots/triangles represent the average value for each neuron, Cntnap2+/+ n=17 cells from 9 mice, Cntnap2+/+ + α-DTX n=17 cells from 9 mice.

(D) Quantification (mean ± SEM) of the rheobase current in Cntnap2+/+ iSPNs. Dots/triangles represent the rheobase current for each neuron. n is the same as in panels B-C. p=0.5267, two-tailed unpaired t test.

(E-H) Quantification (mean ± SEM) of the Cntnap2+/+ iSPN fast afterhyperpolarization (E), *p=0.0238, Mann-Whitney test; resting membrane potential (F), p=0.6832, Mann-Whitney test; AP height (G), p=0.6448, two-tailed unpaired t test; and AP width at half max (H), p=0.3070, two-tailed unpaired t test. Dots/triangles represent the value for the first spike evoked at rheobase for each neuron. n is the same as in panels B-C.

(I) Quantification (mean ± SEM) of Cntnap2+/+ iSPN AP latency to first spike, plotted by number of APs fired in the current pulse. Shaded area around line represents SEM. Mixed-effects model with Geisser-Greenhouse correction p values are shown; d x APs F (17, 82) = 2.375, drug F (1, 32) = 6.104, APs F (17.00, 82.00) = 40.22.

(J) Phase plane plot of a single AP evoked in Cntnap2−/− iSPNs in the absence (solid blue) or presence (dotted blue) of α-Dendrotoxin (α-DTX).

(K-L) Quantification (mean ± SEM) of the Cntnap2−/− iSPNs membrane resistance (K), p=0.1746, two-tailed unpaired t test; and membrane capacitance (L), p=0.7511, two-tailed unpaired t test. Squares/triangles represent the average value for each neuron, Cntnap2−/− n=18 cells from 10 mice, Cntnap2−/− + α-DTX n=19 cells from 10 mice.

(M) Quantification (mean ± SEM) of the rheobase current in Cntnap2−/− iSPNs. Squares/triangles represent the rheobase current for each neuron. n is the same as in panels K-L. p=0.6589, Mann-Whitney test.

(N-Q) Quantification (mean ± SEM) of the Cntnap2−/− iSPN fast afterhyperpolarization (N), p=0.0538, two-tailed unpaired t test; resting membrane potential (O), p=0.9461, two-tailed unpaired t test; AP height (P), p=0.5920, two-tailed unpaired t test; and AP width at half max (Q), *p=0.0335, two-tailed unpaired t test, in the absence or presence of α-DTX. Squares/triangles represent the value for the first spike evoked at rheobase for each neuron. n is the same as in panels K-L.

(R) Quantification (mean ± SEM) of Cntnap2−/− iSPN AP latency to first spike in the absence or presence of α-DTX, plotted by number of APs fired in the current pulse. Shaded area around line represents SEM. Mixed-effects model with Geisser-Greenhouse correction p values are shown; d x APs F (17, 68) = 3.181, drug F (1, 35) = 20.66, APs F (17.00, 68.00) = 82.73.

DeepLabCut and Keypoint-MoSeq analysis of Cntnap2−/− mice.

(A) Quantification (mean ± SEM) of the total distance traveled in the open field measured by DeepLabCut, p=0.8201, Mann-Whitney test; Cntnap2+/+ n=13 mice and Cntnap2−/− n=11 mice.

(B) Quantification (mean ± SEM) of the number of grooming bouts in the open field assessed by Keypoint-MoSeq, calculated as the number of instances of all grooming syllables. Cntnap2+/+ n=13 mice and Cntnap2−/− n =11 mice, *p = 0.0352, Mann-Whitney test.

(C) Map depicting the difference in transitions between syllables (top 25 most frequent syllables) for Cntnap2+/+ and Cntnap2−/− mice (represented as Cntnap2−/− minus Cntnap2+/+). A red circle around a syllable number indicates increased usage of that syllable in Cntnap2−/− mice, a blue circle indicates decreased usage of that syllable in Cntnap2−/− mice. A red line between syllable numbers indicates increased usage of that syllable transition in Cntnap2−/− mice, a blue line indicates decreased usage of that syllable transition in Cntnap2−/− mice. The thickness of the line scales with the size of the difference and the size of the circle surrounding the syllable number scales with usage of the syllable.

Experimental procedures

Mice

All animal procedures were conducted in accordance with protocols approved by the University of California, Berkeley Institutional Animal Care and Use Committee (IACUC) and Office of Laboratory Animal Care (OLAC) (AUP-2016-04-8684-3). Cntnap2−/− mice and littermate Cntnap2+/+ controls with the following alleles were utilized for each experiment.

Mice were group housed on a 12 h light/dark cycle (dark cycle 9:00 AM – 9:00 PM) and given ad libitum access to standard rodent chow and water. Both male and female animals were used for experimentation. The ages, sexes, and numbers of mice used for each experiment are indicated in the respective method details and figure legends. All mice used for experiments were heterozygous or hemizygous for the Drd1a-tdTomato, Thy1-ChR2-YFP, PV-Cre, Ai32, or Ai9 transgenes to avoid potential physiological or behavioral alterations.

Electrophysiology

Mice (P50-60) were briefly anesthetized with isoflurane and perfused transcardially with ice-cold ACSF (pH = 7.4) containing (in mM): 127 NaCl, 25 NaHCO3, 1.25 NaH2PO4, 2.5 KCl, 1 MgCl2, 2 CaCl2, and 25 glucose, bubbled continuously with carbogen (95% O2 and 5% CO2). Brains were rapidly removed and coronal slices (275 μm) were cut on a VT1000S vibratome (Leica) in oxygenated ice-cold choline-based external solution (pH = 7.8) containing (in mM): 110 choline chloride, 25 NaHCO3, 1.25 NaHPO4, 2.5 KCl, 7 MgCl2, 0.5 CaCl2, 25 glucose, 11.6 sodium ascorbate, and 3.1 sodium pyruvate. Slices were recovered in ACSF at 36°C for 15 min and then kept at room temperature (RT) before recording. Recordings were made with a MultiClamp 700B amplifier (Molecular Devices) at RT using 3-5 MOhm glass patch electrodes (Sutter, #BF150-86-7.5). Data were acquired using ScanImage software, written and maintained by Dr. Bernardo Sabatini (https://github.com/bernardosabatini/SabalabAcq). Traces were analyzed in Igor Pro (Wavemetrics). Recordings with a series resistance > 25 MOhms or holding current above 200 pA were rejected. Passive properties were calculated using the double exponential curve fit of the average of five –5 mV, 100 ms long pulse steps applied at the beginning of every experiment.

Current-clamp recordings

Current clamp recordings were made using a potassium-based internal solution (pH = 7.4) containing (in mM): 135 KMeSO4, 5 KCl, 5 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 10 phosphocreatine, and 1 EGTA. For corticostriatal excitability experiments, optogenetic stimulation consisted of a full-field pulse of blue light (470 nm, 0.5 ms pulse width, CoolLED) through a 63x objective (Olympus, LUMPLFLN60XW). Light power was linear over the range of intensities tested. No synaptic blockers were included. For intrinsic excitability experiments (SPN, PV interneuron, and SPN + α-DTX experiments), NBQX (10 μM, Tocris, #1044), CPP (10 μM, Tocris, #0247) and picrotoxin (50 μM, Abcam, #120315) were added to the external solution to block synaptic transmission. For Kv1.2 inhibition experiments, α-DTX (100nM, Alomone Labs, #D-350) was added to the external solution. Control recordings in the absence of α-DTX were performed on slices prior to drug application or on fresh slices after drug washout in alternating order across recording days. Bovine serum albumin (BSA, 0.005%, Sigma, #A7030) was included in both control and α-DTX-containing external solutions to minimize nonspecific binding. One second depolarizing current steps were applied to induce action potentials. No holding current was applied to the membrane.

Voltage-clamp recordings

Voltage-clamp recordings were made using a cesium-based internal solution (pH = 7.4) containing (in mM): 120 CsMeSO4, 15 CsCl, 10 TEA-Cl, 8 NaCl, 10 HEPES, 1 EGTA, 5 QX-314, 4 Mg-ATP, and 0.3 Na-GTP. Recordings were acquired with the amplifier Bessel filter set at 3 kHz. Corticostriatal synaptic stimulation experiments to measure evoked EPSCs were performed in picrotoxin (50 μM), and optogenetic stimulation consisted of a full-field pulse of blue light (470 nm, 0.15 ms pulse width) through a 63x objective. To record AMPAR-mediated EPSCs, cells were held at –70mV; to record NMDAR-mediated EPSCs, cells were held at +40mV. Synaptic stimulation experiments to measure evoked IPSCs were performed in NBQX (10 μM) and CPP (10 μM). For electrically evoked IPSCs, a concentric bipolar stimulating electrode (FHC, #30214) was placed in dorsal striatum, roughly 200 μm medial to the recording site in dorsolateral striatum, and a 0.15 ms stimulus was applied. For PV-interneuron optically evoked IPSCs, a full-field pulse of blue light (470 nm, 0.15 ms pulse width) was applied through a 63x objective at the recording site. All evoked IPSC experiments were recorded with cells held at –70mV.

Dendritic imaging and spine analysis

Neonatal (P1-3) Cntnap2−/−;D1-tdT and Cntnap2+/+;D1-tdT mice were cryoanesthetized and injected bilaterally with 200 nL AAV1.hSyn.eGFP.WPRE.bGH (Penn Vector Core, #p1696 (Keaveney et al., 2018)), diluted 1:75 in saline to achieve sparse transduction. Injections were targeted to the dorsal striatum, with coordinates approximately 1.3 mm lateral to midline, 2.0 mm posterior to bregma, and 1.5 mm ventral to the head surface. At P50-60, mice were anesthetized with isoflurane and transcardial perfusion was performed with 10 mL of 1x PBS followed by 10 mL of ice cold 4% PFA (EMS, #15710-S) in 1x PBS. Brains were post-fixed in 4% PFA in 1x PBS overnight at 4° C. 80 μm coronal sections were made using a freezing microtome (American Optical, AO 860) and stored in 1x PBS at 4° C. Sections were blocked for 1 hour at RT in BlockAid (ThermoFisher, #B10710) and incubated for 48 hours with gentle shaking at 4° C with antibodies against GFP (1:2500, Abcam, #13970) and RFP (1:1000, Rockland (VWR, #600-401-379) diluted in PBS-Tx (1x PBS with 0.25% Triton X-100 (Sigma, #T8787). Sections were washed 3 x 10 min in PBS-Tx and incubated with gentle shaking for 1 hour at RT with Alexa Fluor 488 and 546 secondary antibodies (1:500, Invitrogen, #A11039, #A11035). Sections were washed 3 x 10 min in 1x PBS and mounted onto SuperFrost slides (VWR, #48311-703) using VECTASHIELD HardSet Antifade Mounting Medium (Vector Laboratories, #H-1400-10). Z stack images of individual dendrites were taken on a confocal microscope (Olympus FLUOVIEW FV3000) with a 60x oil immersion objective (Olympus #1-U2B832) at 2.5x zoom with a step size of 0.4 μm and deconvoluted using Olympus CellSens software. To quantify spine density, dendrites and spines were reconstructed using the FilamentTracer module in Imaris software (Oxford Instruments). The spine density of each dendrite was calculated using Imaris. Dendrites analyzed varied in total length, but excluded the most proximal and distal portions of the dendrite.

Brain sectioning and immunohistochemistry

Adult mice were perfused as above and brains were post-fixed with 4% paraformaldehyde overnight, then sectioned coronally at 30 μm. For immunohistochemistry, individual wells of sections were washed for 3 x 5 min with 1x PBS, then blocked for 1 hour at RT with BlockAid blocking solution. Primary antibodies diluted in PBS-Tx were added and tissue was incubated for 48 hours with gentle shaking at 4° C. Sections were then washed 3 x 10 min with PBS-Tx. Secondary antibodies diluted 1:500 in PBS-Tx were added and incubated with gentle shaking for 1 hour at RT. Sections were washed 3 x 10 min in 1x PBS. Sections were mounted onto SuperFrost slides (VWR, #48311-703) and coverslipped with VECTASHIELD HardSet with DAPI (Vector Laboratories, #H-1500-10) or VECTASHIELD HardSet Antifade Mounting Medium (Vector Laboratories, #H-1400-10). The following antibodies were used: mouse anti-PV (1:1000, Sigma, #P3088), rabbit anti-PV (1:1000, Abcam, #11427), anti-RFP (1:500, Rockland, #600-401-379), Alexa Fluor 405, 488 and 546 conjugated secondary antibodies (1:500, Invitrogen, #A-31553, #A-11001, #A-11003, and #A-11035).

PV cell counting

To count PV+ interneurons, Z-stack images of immunostained striatal sections were taken on a confocal microscope (Olympus FLUOVIEW FV3000) with a 10x or 20x objective (Olympus # 1-U2B824 or Olympus # 1-U2B825) and step size of 2 μm. For quantification, image stacks were Z-projected to maximum intensity using Fiji (ImageJ) and cropped to a 400 μm x 400 μm image in anatomically matched sections of the DLS. All PV-expressing cells within this region were counted using the ROI manager tool in ImageJ. Designation of a cell as PV positive was determined by the experimenter and consistently maintained across animals. Experimenter was blind to genotype and ROIs were made on the DAPI channel to avoid selecting regions based on PV expression. To quantify bulk PV fluorescence, ROIs were manually defined in ImageJ using the Freehand tool to cover as much of the DLS as possible, and mean fluorescence intensity was measured. To quantify individual cell PV fluorescence, ROIs were manually defined around every PV positive cell in the previously drawn DLS ROI using the Freehand tool, and mean fluorescence intensity was measured.

Western Blot

Adult mice (P48-55) were deeply anesthetized with isoflurane and decapitated. Brains were rapidly dissected and 1.5 mm dorsal striatum punches (Biopunch, Ted Pella, #15111-15) were collected from both hemispheres, flash-frozen in liquid nitrogen, and stored at −80° C. On the day of analysis, frozen samples were sonicated (QSonica Q55) until homogenized in 200 μl lysis buffer containing 1% SDS in 1x PBS with Halt phosphatase inhibitor cocktail (Thermo Fisher Scientific, #PI78420) and Complete mini EDTA-free protease inhibitor cocktail (Roche, #4693159001). Sample homogenates were then boiled on a heat block at 95° C for 5 min and allowed to cool to RT. Total protein content was determined using a BCA assay (Thermo Fisher Scientific, #23227). Following the BCA assay, protein homogenates were mixed with 4x Laemmli sample buffer (BioRad, #161-0747). 12.5μg of total protein per sample were then loaded onto 12% Criterion TGX gels (BioRad, #5671044) and run at 65 V. Proteins were transferred to a PVDF membrane (BioRad, #1620177) at 11 V for 14 hours at 4° C using the BioRad Criterion Blotter (BioRad, #1704070). Membranes (BioRad, #1620177) were briefly reactivated in methanol and rinsed in water 3x. After rinsing, membranes were blocked in 5% milk in 1x TBS with 1% Tween (TBS-Tween) for 1 hour at RT before being incubated with primary antibodies diluted in 5% milk in TBS-Tween overnight at 4° C. The following day, after 3 x 10 min washes with TBS-Tween, membranes were incubated with secondary antibodies for 1 hour at RT. Following 6 × 10 min washes, membranes were incubated with chemiluminescence substrate (PerkinElmer #NEL105001EA) for 1 min and exposed to Amersham Hyperfilm ECL (VWR, #95017-661).

Bands were quantified by densitometry using ImageJ software. GAPDH was used to normalize protein content and data are expressed as a percentage of control within a given experiment. The following antibodies were used: anti-Caspr2 (1:5000, Abcam, #153856), anti-PV (1:2500, Abcam, #11427), anti-GAPDH (1:5000, Cell Signaling, #51745S), and anti-rabbit goat HRP conjugate (1:5000, BioRad, #1705046).

In situ hybridization

Fluorescent in situ hybridization was performed to quantify Pvalb mRNA expression in the striatum of Cntnap2+/+ and Cntnap2−/− mice. Mice were briefly anesthetized with isoflurane and brains were harvested, flash-frozen in OCT mounting medium (Thermo Fisher Scientific, #23-730-571) on dry ice and stored at –80° C for up to 6 months. 16 µm sections were collected using a cryostat (Thermo Fisher Scientific, Microm HM 550), mounted directly onto Superfrost Plus glass slides (VWR, #48311-703) and stored at –80° C for up to 6 months. In situ hybridization was performed according to the protocols provided with the RNAscope Multiplex Fluorescent Reagent Kit (ACD, #323100). Drd1a mRNA was visualized with a probe in channel 2 (ACD, #406491-C2) and Pvalb mRNA in channel 3 (ACD, #421931-C3). After incubation, sections were secured on slides using ProLong Gold Antifade Mountant with DAPI (Invitrogen, P36935) and 60 x 24 mm rectangular glass coverslips (VWR, #16004-096). Sections were imaged on an Olympus FluoView 3000 confocal microscope using a 10x objective with 1.5x zoom and a step size of 2 µm. Pvalb-expressing cells were quantified across the entire striatum using the ROI manager tool in ImageJ. A cell was considered Pvalb positive if over 50% of the cell contained fluorescent puncta when compared to the DAPI channel. Experimenter was blind to genotype.

Behavioral analysis

All behavior studies were carried out in the dark phase of the light cycle under red lights (open field) or white lights (marble burying, holeboard, rotarod, and four choice reversal learning). Mice were habituated to the behavior testing room for at least 30 min prior to testing. Mice were given at least one day between different tests. All behavior equipment was cleaned between each trial and mouse with 70% ethanol and rinsed in diluted soap followed by water at the end of the day. If male and female mice were to be tested on the same day, male mice were run first then returned to the housing room, after which all equipment was thoroughly cleaned prior to bringing in female mice for habituation. Behavioral tests were performed with young adult male and female mice (7-11 weeks old). The experimenter was blind to genotype throughout the testing and scoring procedures.

Open field assay

Exploratory behavior in a novel environment and general locomotor activity were assessed by a 60 min session in an open field chamber (40 cm L x 40 cm W x 34 cm H) made of transparent plexiglass. Horizontal infrared photobeams (Stoelting, 60001-02A) were positioned to detect rearing. The mouse was placed in the bottom right-hand corner of the arena and behavior was recorded using an overhead camera and analyzed using ANY-maze software (Stoelting). An observer manually scored self-grooming behavior during the first 20 minutes of the test. A grooming bout was defined as an unbroken series of grooming movements, including licking of body, paws, or tail, as well as licking of forepaws followed by rubbing of face with paws.

Open field assay with DeepLabCut Keypoint-MoSeq analysis

Mice were placed in the open field arena and video recorded with a monochrome camera (FLIR Grasshopper 3, GS3-U3-41C6NIR-C) and a 16 mm wide angle lens (Kowa, LM16HC) placed above the arena from a height of 50 cm. To extract the body part (keypoint) coordinates from the video recordings, DeepLabCut (DLC) 2.3.4 (Mathis, et al. 2018; Nath, et al. 2019) was used. Fourteen body parts including nose, head, left ear, right ear, left forelimb, right forelimb, spine 1, spine 2, spine 3, left hindlimb, right hindlimb, tail 1, tail 2, and tail 3 were manually labeled on a small subset of the video frames. A DLC model was then trained using the annotated frames to label those 14 body parts for all videos recorded. The total distance traveled, and number of center entries were calculated using the coordinate of bodypart tail 1.

Discrete behavior syllables were extracted using Keypoint-MoSeq 0.4.4 (Weinreb, et al. 2023). Syllable usage and transition data were obtained using built-in functions of the Keypoint-MoSeq package. Decoding and entropy analysis were performed using customized Python 3.9 script. Code available upon request in Github. Entropy was calculated using the following equation, where ui denotes the frequency of the syllable i and pi,jdenotes the transition probability from syllable i to syllable j.:

Marble burying assay

The marble burying assay was used to test for repetitive behavior. 20 black marbles were arranged in an orderly 4 x 5 grid on top of 5 cm of clean corn cob bedding in a standard mouse cage. Overhead room lights were on and white noise was played to induce mild stress. Mice were placed in the cage with the marbles for 30 minutes. The number of unburied marbles (>50% exposed) was recorded after the session.

Holeboard assay

The holeboard assay was used to measure exploratory and repetitive behavior. The holeboard apparatus consisted of a smooth, flat, opaque gray plastic platform, suspended 10 cm from the base by four plastic pegs in each corner. The board contained 16 evenly spaced 2 cm diameter holes and was surrounded by a 30 cm high clear plastic square encasing. During testing, mice were placed into the center of the holeboard. Mice explored the board for 10 minutes while video was recorded from both an above and side-view camera. Videos were used post-hoc to manually count and map the number of nose pokes made during the task. Nose pokes were defined as the mouse’s nose passing through the board barrier when viewed through the side-view camera.

Accelerating rotarod assay

The accelerating rotarod test was used to examine motor coordination and learning.

Mice were trained on a rotarod apparatus (Ugo Basile, #47650) for four consecutive days. Three trials were completed per day with a 5 min break between trials. The rotarod was accelerated from 5-40 revolutions per minute (rpm) over 300 s for trials 1-6 (days 1 and 2), and from 10-80 rpm over 300 s for trials 7-12 (days 3 and 4). On the first testing day, mice were first acclimated to the apparatus by being placed on the rotarod rotating at a constant 5 rpm for 60 s and returned to their home cage for 5 min prior to starting trial 1. Latency to fall, or to rotate off the top of the rotarod barrel, was measured by the rotarod stop-trigger timer.

Four choice odor-based reversal learning test

The four-choice odor-based reversal learning test was used to assess learning and cognitive flexibility. Animals were food restricted for 6 days in total, with unrestricted access to drinking water, and maintained at 90-95% of ad lib feeding body weight. Food was given at the end of the day once testing was completed. Food restriction and introduction to Froot Loop cereal piece (Kellogg’s, Battle Creek, MI) began 48 hours before pre-training. The four-choice test was performed in a custom-made square box (30.5 cm L × 30.5 cm W × 23 cm H) constructed of clear acrylic. Four internal walls 7.6 cm wide partially divided the arena into four quadrants. A 15.2 cm diameter removable cylinder fit in the center of the maze and was lowered between trials (after a digging response) to isolate the mouse from the rest of the maze. Odor stimuli were presented mixed with wood shavings in white ceramic pots measuring 7.3 cm in diameter and 4.5 cm deep. All pots were sham baited with a piece of Froot Loop cereal secured underneath a mesh screen at the bottom of the pot. This was to prevent mice from using the odor of the Froot Loop to guide their choice. The apparatus was cleaned with 2.5% acetic acid followed by water and the pots were cleaned with 70% ethanol followed by water between mice. The apparatus was cleaned with diluted soap and water at the end of each testing day.

On the first habituation day of pre-training (day 1), animals were allowed to freely explore the testing arena for 30 min and consume small pieces of Froot Loops placed inside empty pots positioned in each of the four corners. On the second shaping day of pre-training (day 2), mice learned to dig to find cereal pieces buried in unscented coarse pine wood shavings (Harts Mountain Corporation, Secaucus, NJ). A single pot was used and increasing amounts of unscented wood shavings were used to cover each subsequent cereal reward. The quadrant containing the pot was alternated on each trial and all quadrants were rewarded equally. Trials were untimed and consisted of (in order): two trials with no shavings, two trials with a dusting of shavings, two trials with the pot a quarter full, two trials with the pot half full, and four trials with the cereal piece completely buried by shavings. The mouse was manually returned to the center cylinder between trials.

On the days for odor discrimination (day 3, acquisition) and reversal (day 4), wood shavings were freshly scented on the day of testing. Anise extract (McCormick, Hunt Valley, MD) was used undiluted at 0.02 ml/g of shavings. Clove, litsea, and eucalyptus oils (San Francisco Massage Supply Co., San Francisco, CA) were diluted 1:10 in mineral oil and mixed at 0.02 ml/g of shavings. Thymol (thyme; Alfa Aesar, A14563) was diluted 1:20 in 50% ethanol and mixed at 0.01 ml/g of shavings. During the discrimination phase (day 3), mice had to discriminate between four pots with four different odors and learn which one contained a buried food reward. Each trial began with the mouse confined to the start cylinder. Once the cylinder was lifted, timing began, and the mouse could freely explore the arena until it chose to dig in a pot. Digging was defined as purposefully moving the shavings with both front paws. A trial was terminated if no choice was made within 3 min and recorded as omission. Criterion was met when the animal completed eight out of ten consecutive trials correctly. The spatial location of the odors was shuffled on each trial. The rewarded odor during acquisition was anise.

The first four odor choices made during acquisition were analyzed to determine innate odor preference by the percentage of choices for a given odor: Cntnap2+/+ mice: 60% thyme, 25% anise, 12.5% clove, and 2.5% litsea. Cntnap2−/− mice: 47.5% thyme, 45% anise, 7.5% clove, 0% litsea. We note that both Cntnap2+/+ and Cntnap2−/− mice exhibited the strongest innate preference for thyme, an unrewarded odor. There were no significant differences in innate odor preference.

The reversal phase of the task was carried out on day 4. Mice first performed the task with the same rewarded odor as the discrimination day to ensure they learned and remembered the task. After reaching criterion on recall (eight out of ten consecutive trials correct), the rewarded odor was switched, and mice underwent a reversal learning test in which a previously unrewarded odor (clove) was rewarded. A novel odor (eucalyptus) was also introduced, which replaced thyme. Perseverative errors were choices to dig in the previously rewarded odor that was no longer rewarded. Regressive errors were choosing the previously rewarded odor after the first correct choice of the newly rewarded odor. Novel errors were choices to dig in the pot with the newly introduced odor (eucalyptus). Irrelevant errors were choices to dig in the pot that had never been rewarded (litsea). Omissions were trials in which the mouse failed to make a digging choice within 3 min from the start of the trial. Total errors were the sum of perseverative, regressive, irrelevant, novel, and omission errors. Criterion was met when the mouse completed eight out of ten consecutive trials correctly. The spatial location of the odors was shuffled on each trial.

Quantification and statistical analysis

Experiments were designed to compare the main effect of genotype. The sample sizes were based on prior studies and are indicated in the figure legend for each experiment. Whenever possible, quantification and analyses were performed blind to genotype. GraphPad Prism version 10 was used to perform statistical analyses. The statistical tests and outcomes for each experiment are indicated in the respective figure legend. Two-tailed unpaired t tests were used for comparisons between two groups. For data that did not pass the D’Agostino & Pearson normality test, a Mann-Whitney test was used. Two-way ANOVAs or mixed effects models were used to compare differences between groups for experiments with two independent variables. Statistical significance was defined in the figure panels as follows: *p < 0.05, **p < 0.01, ***p < 0.001.